---
title: "Field and DePlata-Peterson 2026"
format: html
editor: visual
---

# Research compendium for Field and DePlata-Peterson (2026)

This R-based research compendium supports the analyses from *Absolute Precipitation as a Driver of Re-settlement in the northern US Southwest*, which shows that that local measures of absolute and directional change in precipitation are strongly correlated with settlement histories; however, absolute precipitation change is the most closely linked as an explanatory driver of near-term future construction activity.

updating..

View the raw source code and output at: provide link

**When using the code included in this research compendium, please cite *all* of the following:**

> Field, S and M. DePlata-Peterson. 2026. *Variability as a Driver of Re-settlement in the northern US Southwest*. Journal of Archaeological Science, In Prep.
>
> Field, S and M. DePlata-Peterson. 2026. *Research Compendium for Variability as a Driver of Re-settlement in the northern US Southwest*., Version 1.0.0. Zenodo. \[DOI\]

## 1.0 Set Up Environment

```{r setup, include=FALSE}
#| echo: false
knitr::opts_knit$set(root.dir = "C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/PROJECT/CLIMATE_VARIABILITY_REDUX")
```

```{r}
#| echo: false
packages <-c('terra','sf','ggplot2','tidyterra','dplyr','tidyr','gifski','gganimate','ggforce','ggpubr','scales','lmtest','tseries','FedData')

for(p in packages) if(p %in% rownames(installed.packages()) == F) { install.packages(p) }
for(p in packages) suppressPackageStartupMessages(library(p,quietly=T,character.only=T))

theme_set(theme_bw())
```

## 2.0 Import Study Area Extent and Paleoclimate Raster

`study_area`: The study area is derived as the latitude and longitude of extent of sites used in the study (see Section 3.0). This extent can be downloaded as a shapefile [here](https://github.com/sfield2/Field_and_DePlata-Peterson2026/tree/main/Data).

`four_corners`: Shapefiles for the four corners states are used for plotting of supplemental figures later on. These data can be downloaded from the US Census [here.](https://www.census.gov/geographies/mapping-files/time-series/geo/carto-boundary-file.html)

```{r}
#| echo: false
study_extent <- st_read("./DATA/study_extent.shp")%>%
  st_transform(.,crs=4326)

four_corners <- st_read("C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/DATA_GENERAL/SPATIAL/NORTH_AMERICA_REGIONS/SOUTHWEST/STATE_BOUNDARIES/four_corners_states.shp")%>%
  st_transform(.,crs=4326)%>%
  st_intersection(.,study_extent)
```

`paleoclimate_raster`: The paleoclimate raster is from PaleoCAR, a high resolution 2000 year reconstruction of annual precipitation and temperature based on a correlation-adjusted regression of tree-rings and contemporary PRISM models (Bocinsky et al. 2016). These data can be accessed several ways:

1.  Access archives directly as NetCDF files as part of the NOAA National Centers for Environmental Information Paleoclimate archive at: [DOI](https://doi.org/10.25921/8ctk-8v26).
2.  Access via SKOPE (associated github repository [here](https://github.com/bocinsky/paleocar_v2)). The scripts in this repository download the archived tiles, mosaic and write them as cloud-optimized GeoTIFFs, then uploads them to an [Amazon AWS S3 bucket](https://skope.s3.us-west-2.amazonaws.com/index.html), allowing them to queried remotely. The approach detailed below is adopted from <https://github.com/bocinsky/paleocar_v2>.
    1.  **Note:** accessing via SKOPE requires transforming to WGS 84 CRS (EPSG:4326)

Both options are provided below:

```{r}
#| echo: false
# OPTION 1: 
# r <- terra::rast("./DATA/CLIMATE/PPT.tif")

# OPTION 2:
study_extent <- study_extent%>%
  st_transform(.,4326)

r <- terra::rast("https://skope.s3.us-west-2.amazonaws.com/paleocar_v2/ppt_wateryear/prediction_scaled.tif", vsi = TRUE) %>%
  terra::subset(subset = 600:1351) %>%
  terra::crop(terra::vect(study_extent),mask = TRUE)

# check projection and reproject if necessary
# crs(r)
# r<-project(r,"epsg:4326")

# OPTIONAL: write raster to local
writeRaster(r,"./DATA/CLIMATE/PPT_600_1351.tif")
```

### 2.1 Calculate Base Precipitation Variability Raster

Calculate year-to-year change as a measure of temporal variability. This process calculates the difference from every year (2-2000 C.E.) and the previous year (1-1999 C.E.) for each cell.

```{r}
#| echo: false
r_var <- subset(r, 2:nlyr(r)) - subset(r, 1:(nlyr(r)-1))

writeRaster(r_var,"./DATA/CLIMATE/PPT_annualvariability_600_1351.tif")
```

**NOTE:** After variability assessment is computed/exported, you can read in directly to avoid re-running.

```{r}
#| echo: false

# OPTIONAL: 
# r <- terra::rast("./DATA/CLIMATE/PPT_600_1351.tif")
# r_var <- terra::rast("./DATA/CLIMATE/PPT_annualvariability_600_1351.tif")
```

## 3.0 Import Site Data

### 3.1 Chaco Great Houses Site Data

Settlement data of Chaco Great House communities comes from two sources:

1.  `chaco_greathouses`: This is the *Greater Chaco Landscapes Great House Communities Integration Dataset*. **Note:** A version of these data were published by Van Dyke et al. (2016, [DOI](https://doi.org/10.7183/0002-7316.81.2.205)). Access to this data can be requested [here](https://doi.org/10.6067/XCV8BV7JJ4). The database contains 365 sites, including:
    1.  15 great houses in the Chaco Core, 243 great house outliers, 11 herraduras, 46 shrines, 21 stone circles, 9 crescents, and 20 important landforms.
    2.  The full citation for the database is: Heitman, Carrie C., Ruth Van Dyke, Matthew Peebles, and Kyle Boczinsky. 2016. Greater Chaco Landscapes Great House Communities GIS Integration Dataset. GIS dataset submitted to the National Park Service in partial fulfillment of Rocky Mountain Cooperative Ecosystems Study Unit Task Agreement Number: P14AC01703, Project #: UCOB-109.
2.  `outlier_database`: This is the Chacoan Outlier database, synthesized by Kantner et al. at the "Chaco World Conference" in 1999. The database was maintained by Kantner (2001-2007), receiving contributions from Paul Reed, R.G. Matson, Bill Lipe, Mark Varien, Scott Ortman, and Susan Ryan. The non-spatial data attributes are publicly accessible at the Chaco Research Archive, and can be accessed [here](https://www.chacoarchive.org/cra/outlier-database/).

The geodatabase `chaco_greathouses` and the attributes from `outlier_database` are merged to reconstruct occupation at 105 Chaco great house outliers. These data are stored under as the object `gh`.

```{r}
#| echo: false
chaco_greathouses <- st_read("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/VanDykeEtAl2015_WGS84.shp")%>%
  st_transform(.,crs=4326)%>%
  st_intersection(.,study_extent)

outlier_database <-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/SENSITIVE SITE LOCATIONS/OutlierAttributes_Public_CRA.csv")%>%
  select("Site.name","Site.number.","Est..great.house.start.date.","Est..great.house.end.date.")%>%
  setNames(c("SITE_NA","SITE_NUMBE","BEGIN_DATE","END_DATE"))

gh <- merge(chaco_greathouses,outlier_database,by="SITE_NA")%>%
  subset(SITE_T == "GH_OUTLIER") %>%
  filter(!is.na(BEGIN_DATE))%>%
  filter(!is.na(END_DATE))%>%
  subset(BEGIN_DATE != "No Data")%>%
  subset(END_DATE != "No Data")%>%
  select(-c("SITE_NUMBE.x","SITE_NUM_1","PROJECTED_","PROJECTE_1","PROJECTE_2"))%>%
  mutate(END_DATE = as.numeric(END_DATE),
         BEGIN_DATE = as.numeric(BEGIN_DATE),
         occupation_length = END_DATE - BEGIN_DATE)%>%
  subset(occupation_length >= 1)

rm(chaco_greathouses,outlier_database)
```

### 3.2 VEP II North Study Area Site Data

Settlement data from the Village Ecodynamics Project (VEP) II north study area come from Schwindt et al. (2016: [DOI](http://doi.org/10.7183/0002-7316.81.1.74)), which includes additions to an earlier (VEP I) database (Ortman et al. 2007: [DOI](https://doi.org/10.2307/40035813)).

1.  `vepii_sites`: The database used in this study is a limited version of the VEP II north database, which before filtering and subsetting contains 9286 site records. The data was subset to include only record site types with characteristics common to residential sites.
2.  `vep_dates`:To aid in later stages creating a complete occupation time series for all sites in the VEP database (see section 5.1.2), the date ranges for each occupational phase are imported. These date ranges can be accessed [here](https://github.com/sfield2/Field_and_DePlata-Peterson2026/tree/main/Data) or imported following the code below.

```{r}
#| echo: false
vepii_sites<-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/SENSITIVE SITE LOCATIONSdb_vepii_n.csv")%>%
  subset(tlkp.Site.Function_DescGen=="Habitation")%>%
  subset(N.roomblocks >= 1)%>%
  subset(recordtypedesc=="Habitation" | recordtypedesc == "Great House"| recordtypedesc=="Oversized pitstructure" | recordtypedesc == "Great Kiva"| recordtypedesc=="Multi-wall Structure" | recordtypedesc == "Tower-Kiva") %>%
  #subset(Dataquality != 3)%>%
  st_as_sf(.,coords=c("UTMEast","UTMNorth"),crs=26912)%>%
  st_transform(.,crs=4326)%>%
  st_intersection(.,study_extent)%>%
  mutate(UNIQ_ID = 1:nrow(.))

vep_dates <- read.csv("https://raw.githubusercontent.com/sfield2/Field_and_DePlata-Peterson2026/refs/heads/main/Data/VEP_dates.csv")
```

## 4.0 Calculate Climate Conditions

In this study, five climate measures are assessed, which can are divided into three primary classes:

1.  **Precipitation availability:**
    1.  `mean_precipitation` - this is the mean annual precipitation within each catchment
2.  **Directional change:**
    1.  `temporal_variability` - this is the mean year-to-year precipitation variability within each catchment
    2.  `5_yr_zscore` - this measure follows McCool et al. (2025: [DOI](https://doi.org/10.1017/aaq.2025.10107)) and measures "...whether the precipitation during any given year is higher or lower than the prior five-year average and what the magnitude of that change is" (pg. 107). This is done by finding the difference between the mean annual precipitation each year and the average for the previous five years and dividing it by the standard deviation of annual precipitation for the previous five years.
3.  **Absolute change:**
    1.  `spatial_variability` - this is the annual standard deviation within each catchment.

    2.  `spatiotemporal_variability` - this is the standard deviation year-to-year precipitation variability within each catchment. **NOTE:** in most places all of the change is either negative or positive, so usually absolute change is the additive inverse. However, in some places where some cells change positively and some negatively, the absolute change is different.

To calculate the above measures of precipitation availability and change for each site, the following steps were used:

1.  Select a site, apply a geospatial buffer by 2 km to identify approximate local catchment, and catalog site name or ID
2.  Extract the raster data for the buffered area (both annual precipitation and year-to-year precipitation change raster)
3.  Summarize the five climate measures from the associated raster extractions and combine for easier interpretation later.

### 4.1 Demonstrate climate measure calculations

Below is a visual demonstration of the analysis, to aid in understanding the processing steps used to delineate the local catchment for each site and summarize the

```{r}
#| echo: false

# this is a site location from a publicly accessible site in Mesa Verde National Park
site <- data.frame(site_name = "public_site",
                   latitude = 37.24,
                   longitude = -108.50)%>%
  st_as_sf(coords=c("longitude","latitude"),crs=4326)

### 1. Select a site and apply buffer
site_buff <- site[1,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

site_name <- site[1,c("site_name")]%>%
              st_drop_geometry()%>%
              as.character()

### 2. Crop or extract raster data from within buffer extent
# NOTE: In this example raster data is cropped to more easily visualize. However, during the full processing, the data is extracted and transformed into a dataframe immediately
site_precip <- crop(r,site_buff)
site_var <- crop(r_var,site_buff) 


### 3: Summarize the associated metrics.
site_precip_df <- terra::extract(r,site_buff,exact=T)
site_var_df <- terra::extract(r_var,site_buff,exact=T)

mean_precipitation <- site_precip_df%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("mean_precipitation","year"))
  
  spatial_variability <- site_precip_df%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("spatial_variability","year"))
  
  temporal_variability <- site_var_df%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("temporal_variability","year"))

  spatiotemporal_variability <- site_var_df%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("spatiotemporal_variability","year"))
  
  z_scores <- mean_precipitation%>%
    mutate(five_year_av = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=mean,fill=NA),
            five_year_sd = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=sd,fill=NA),precip_5yr_zscore = (mean_precipitation-five_year_av)/five_year_sd)%>%
  select(c("year","precip_5yr_zscore"))
  
summary <- merge(mean_precipitation,spatial_variability,by="year")%>%
  merge(.,temporal_variability,by="year")%>%
  merge(.,spatiotemporal_variability,by="year")%>%
  merge(.,z_scores,by="year")%>%
  gather(.,key="variable",value="value",-year)%>%
  setNames(c("year","variable",site_name))

```

#### 4.1.1 Demonstrate climate measure calculations

```{r}
#| echo: false

p1 <- ggplot(site_precip[[451]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1050`))+
  geom_text(aes(label=round(`1050`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1050",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p2 <- ggplot(site_precip[[452]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1051`))+
  geom_text(aes(label=round(`1051`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1051",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p3 <- ggplot(site_precip[[453]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1052`))+
  geom_text(aes(label=round(`1052`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1052",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p4 <- ggplot(site_precip[[454]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1053`))+
  geom_text(aes(label=round(`1053`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1053",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p5 <- ggplot(site_precip[[455]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1054`))+
  geom_text(aes(label=round(`1054`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1054",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p6 <- ggplot(site_precip[[456]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1055`))+
  geom_text(aes(label=round(`1055`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Year 1055",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

png('./FIGURES/Figure_2a.png',height=8000,width=48000,res=1200)
ggarrange(p1,p2,p3,p4,p5,p6,ncol=6,align="hv")
dev.off()
  


p1 <- ggplot(site_var[[451]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1051`))+
  geom_text(aes(label=round(`1051`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Difference between Year 1051 and 1050",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p2 <- ggplot(site_var[[452]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1052`))+
  geom_text(aes(label=round(`1052`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Difference between Year 1052 and 1051",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p3 <- ggplot(site_var[[453]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1053`))+
  geom_text(aes(label=round(`1053`)),col="white")+
  scale_fill_viridis_c(direction= -1,option="magma")+
  labs(title="Difference between Year 1053 and 1052",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p4 <- ggplot(site_var[[454]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1054`))+
  geom_text(aes(label=round(`1054`)),col="white")+
  scale_fill_viridis_c(direction= -1)+
  labs(title="Difference between Year 1054 and 1053",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p5 <- ggplot(site_var[[455]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1055`))+
  geom_text(aes(label=round(`1055`)),col="white")+
  scale_fill_viridis_c(direction= -1,option="magma")+
  labs(title="Difference between Year 1055 and 1054",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

p6 <- ggplot(site_var[[456]],aes(x=x,y=y))+
  geom_raster(aes(fill=`1056`))+
  geom_text(aes(label=round(`1056`)),col="white")+
  scale_fill_viridis_c(direction= -1,option="magma")+
  labs(title="Difference between Year 1056 and 1055",
       x=NULL,
       y=NULL)+
  theme(legend.position = "none")

png('./FIGURES/Figure_2b.png',height=8000,width=48000,res=1200)
ggarrange(p1,p2,p3,p4,p5,p6,ncol=6,align="hv")
dev.off()
```

### 4.2 Calculate Climate Measures

These analyses demonstrated above (Section 4.1) are iterated across all sites in the `gh` and `vepii_sites` databases. To account for differences in the structure of the `gh` and `vepii_sites` databases, separate loops are executed for each settlement type.

#### 4.2.1 Calculating Climate Measures at each Chaco Great House

```{r}
#| echo: false

### ITERATION FOR CHACO GREAT HOUSES 
for(i in 1:nrow(gh)){
  site <- gh[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- gh[i,c("SITE_NA")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("mean_precipitation","year"))
  
  spatial_variability <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("spatial_variability","year"))
  
  temporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("temporal_variability","year"))

  spatiotemporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("spatiotemporal_variability","year"))
  
  z_scores <- mean_precipitation%>%
    mutate(five_year_av = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=mean,fill=NA),
            five_year_sd = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=sd,fill=NA),precip_5yr_zscore = (mean_precipitation-five_year_av)/five_year_sd)%>%
  select(c("year","precip_5yr_zscore"))

  
  summary <- merge(mean_precipitation,spatial_variability,by="year")%>%
    merge(.,temporal_variability,by="year")%>%
    merge(.,spatiotemporal_variability,by="year")%>%
    merge(.,z_scores,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_greathouses <- summary
  }else {
    summary_all_greathouses <- merge(summary_all_greathouses,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   spatial_variability,
   temporal_variability,
   spatiotemporal_variability)

write.csv(summary_all_greathouses,"./OUTPUT/climatevariability_greathouses.csv", row.names = FALSE)
```

#### 4.2.2 Calculating Climate Measures at each VEP II North Site

```{r}
#| echo: false

### ITERATION FOR VEP II NORTH SITES
for(i in 1:nrow(vepii_sites)){
  site <- vepii_sites[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- vepii_sites[i,c("UNIQ_ID")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  site_var <- terra::extract(r_var,site,exact=T)

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("mean_precipitation","year"))
  
  spatial_variability <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("spatial_variability","year"))
  
  temporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("temporal_variability","year"))

  spatiotemporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("spatiotemporal_variability","year"))

  z_scores <- mean_precipitation%>%
    mutate(five_year_av = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=mean,fill=NA),
            five_year_sd = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=sd,fill=NA),precip_5yr_zscore = (mean_precipitation-five_year_av)/five_year_sd)%>%
  select(c("year","precip_5yr_zscore"))

  
  summary <- merge(mean_precipitation,spatial_variability,by="year")%>%
    merge(.,temporal_variability,by="year")%>%
    merge(.,spatiotemporal_variability,by="year")%>%
    merge(.,z_scores,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_vep <- summary
  }else {
    summary_all_vep <- merge(summary_all_vep,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   spatial_variability,
   temporal_variability,
   spatiotemporal_variability)

write.csv(summary_all_vep,"./OUTPUT/climatevariability_vep_habitations.csv", row.names = FALSE)
```

## 5.0 Integrate Climate and Settlement Data

Climate measures were calculated annually from 600 to 1350 C.E. and tabulated as annual time series. To integrate these climate reconstructions and the settlement data, the site databases need to be transformed into time series.

### 5.1 Create Occupation Time Series

Rather than examining the correspondence between periods of climate and social stress, the occupational timelines for all sites were inverted to measure the number of years until resettlement.

#### 5.1.1 Chaco Great House Occupation Time Series

Chaco great houses used in this analysis did not have multi-component occupations, therefore calculating the years before resettlement were estimated easily by counting back from the occupational `"END_DATE"`.

```{r}
#| echo: false
gh_occupation_timeseries <- gh %>%
  rowwise() %>%
  st_drop_geometry()%>%
  mutate(year = list(BEGIN_DATE:END_DATE)) %>%
  unnest(year)%>%
  mutate(years_until_resettlement=END_DATE - year)%>%
  select(c("SITE_NA","year","years_until_resettlement"))%>%
  setNames(c("ID","year","years_until_resettlement"))

write.csv(gh_occupation_timeseries,"./OUTPUT/greathouses_occupation_timeseries.csv", row.names = FALSE)

```

#### 5.1.2 VEP II Occupation Time Series

Many sites in the `vepii_sites` database were multi-component and thus required separating into their separate components, counting the years before resettlement in each component, and then recombining.

```{r}
#| echo: false
vep_occupation_timeseries <- vepii_sites%>%
  st_drop_geometry()%>%
  select(c("ph6","ph7","ph8","ph9","ph10","ph11","ph12",
           "ph13","ph14","ph15","ph16","ph17","ph18","ph19","UNIQ_ID"))%>%
  gather(phase,value,-UNIQ_ID)%>%
  mutate(value = ifelse(value<0,0,value))%>%
  left_join(.,vep_dates,c("phase"="PHASE"))%>%
  arrange(.,desc(UNIQ_ID),BEGIN_D)%>%
  subset(value>0)%>%
  select(-c("phase","value"))%>%
  setNames(c("ID","BEGIN_DATE","END_DATE"))%>%
  arrange(ID,BEGIN_DATE)%>%
  group_by(ID)%>%
  mutate(continuous = BEGIN_DATE==lag(END_DATE)+1,
         group = cumsum(is.na(continuous)|!continuous))%>%
  group_by(ID,group)%>%
  summarise(BEGIN_DATE=min(BEGIN_DATE),
            END_DATE = max(END_DATE))%>%
  rowwise() %>%
  mutate(year = list(BEGIN_DATE:END_DATE))%>%
  unnest(year)%>%
  mutate(years_until_resettlement=END_DATE - year)%>%
  select(c("ID","year","years_until_resettlement"))


write.csv(vep_occupation_timeseries,"./OUTPUT/vep_occupation_timeseries.csv", row.names = FALSE)
```

### 5.2 Merge Climate and Occupational Time Series

#### 5.2.1 Chaco Great House Climate and Occupation

```{r}
#| echo: false
gh_climate_timeseries<-summary_all_greathouses%>%
  gather(key=ID,value=climate_value,-c("year","variable"))

gh_combined_timeseries <- merge(gh_climate_timeseries,gh_occupation_timeseries,by=c("ID","year"))
```

#### 5.2.2 VEP II North Climate and Occupation

```{r}
#| echo: false
vep_climate_timeseries<-summary_all_vep%>%
  gather(key=ID,value=climate_value,-c("year","variable"))%>%
  mutate(ID=as.numeric(sub("^X","",ID)))

vep_combined_timeseries <- merge(vep_climate_timeseries,vep_occupation_timeseries,by=c("ID","year"))
```

### 5.3 Merge All Data

```{r}
#| echo: false
master_output <- rbind(gh_combined_timeseries,vep_combined_timeseries)

# OPTIONAL: write output to enable importing later without re-running analysis

# write.csv(master_output,"./OUTPUT/master_output.csv", row.names = FALSE)
# master_output <- read.csv("./OUTPUT/master_output.csv")
```

## 6.0 Analyze Settlement Results

### 6.1 Plot all results

```{r}
#| echo: false

master_summary <- master_output%>%
  group_by(variable,years_until_resettlement)%>%
  summarize(mean_value = mean(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_resettlement>10 & years_until_resettlement <670)

master_95 <- master_output%>%
  group_by(variable,years_until_resettlement) %>%
  filter(climate_value >= quantile(climate_value, 0.95, na.rm = TRUE))%>%
  summarize(mean_value = min(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_resettlement>10 & years_until_resettlement <670)

master_05 <- master_output%>%
  group_by(variable,years_until_resettlement) %>%
  filter(climate_value <= quantile(climate_value, 0.05, na.rm = TRUE))%>%
  summarize(mean_value = max(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_resettlement>10 & years_until_resettlement <670)

  
png('./FIGURES/all_variables.png',height=26000,width=20000,res=1200)
ggplot(data=master_output,aes(years_until_resettlement,climate_value))+
  geom_line(aes(group=ID),alpha=0.01,linewidth=2,col="#8AB17D")+
  geom_line(data=master_summary,aes(years_until_resettlement,mean_value),color="grey20",linewidth=0.45)+
    geom_line(data=master_95,aes(years_until_resettlement,mean_value),color="grey50",linewidth=0.25)+
    geom_line(data=master_05,aes(years_until_resettlement,mean_value),color="grey50",linewidth=0.25)+
  labs(x="Years before Resettlement",
       y="Climate Variable (mm)",
       title=NULL)+
    theme_minimal()+
  facet_wrap(~variable,scale="free_y",ncol=1,labeller=as_labeller(c(`mean_precipitation` = "Mean Annual Precipitation",`temporal_variability` = "Year-to-Year Change in Mean Precipitation (Temporal Variability)",`spatial_variability`="Standard Deviation in Annual Precipitation (Spatial Variability)",`spatiotemporal_variability`="Year-to-Year Change in Standard Deviation Precipitation (Spatio-Temporal Variability)",`precip_5yr_zscore`="5-year Z-score")))+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        strip.text=element_text(size=15),
        axis.title = element_text(size=16,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10))
dev.off()
```

### 6.2 Examine Correlations and the Effect of Extreme Conditions

Calculate correlations among years until resettlement and the five climate measure. First, the `vepii_sites` data need to be simplified to make it easier to tabulate the associated settlement attributes (site length of occupation, site size, etc).

Following observation that most climate variables dispersed as resettlement grew nearer, correlations were also run on subsets of each climate measure to see how the more extreme experiences of each condition related to resettlement.

```{r}
#| echo: false
vep_occupation_timeseries_simplified <- vepii_sites%>%
  st_drop_geometry()%>%
  select(c("ph6","ph7","ph8","ph9","ph10","ph11","ph12",
           "ph13","ph14","ph15","ph16","ph17","ph18","ph19","UNIQ_ID"))%>%
  gather(phase,value,-UNIQ_ID)%>%
  mutate(value = ifelse(value<0,0,value))%>%
  left_join(.,vep_dates,c("phase"="PHASE"))%>%
  arrange(.,desc(UNIQ_ID),BEGIN_D)%>%
  subset(value>0)%>%
  select(-c("phase","value"))%>%
  setNames(c("ID","BEGIN_DATE","END_DATE"))%>%
  arrange(ID,BEGIN_DATE)%>%
  group_by(ID)%>%
  mutate(continuous = BEGIN_DATE==lag(END_DATE)+1,
         group = cumsum(is.na(continuous)|!continuous))%>%
  group_by(ID,group)%>%
  summarise(BEGIN_DATE=min(BEGIN_DATE),
            END_DATE = max(END_DATE))%>%
  mutate(occupation_length = END_DATE-BEGIN_DATE) 
```

Subset the `master_output` data to make it easier to calculate conditions-specific values above or below percentile thresholds.

```{r}
#| echo: false
master_mean <- subset(master_output, variable == "mean_precipitation")
master_std <- subset(master_output, variable == "spatial_variability")
change_mean <- subset(master_output, variable == "temporal_variability")
change_std <- subset(master_output, variable == "spatiotemporal_variability")
z_score <- subset(master_output, variable == "precip_5yr_zscore")
```

As a demonstration of increased variability as years until resettlement grow nearer, look at the ranges in mean annual precipitation 5 years before resettlement and 400 years before resettlement.

```{r}
#| echo: false

change_mean_5before <- subset(change_mean,years_until_resettlement == 5)
change_mean_400before <- subset(change_mean,years_until_resettlement ==400)

print("The range in mean annual precipitation 400 years before resettlement is:")
range(change_mean_400before$climate_value)

print("The range in mean annual precipitation 5 years before resettlement is:")
range(change_mean_5before$climate_value)

```

#### 6.2.1 Examine Correlations for all values at lowest percentiles

For all climate measures, values at each percentile from ≤ P~49~ to ≤ P~1~ were identified and correlations were calculated. These percentiles are especially important for correctly measuring the most extreme experiences of precipitation availability and directional precipitation change.

```{r}
#| echo: false

mean_df_low <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.01, 0.49, by = 0.01))

std_df_low <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.01, 0.49, by = 0.01))

mean_change_df_low <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.01, 0.49, by = 0.01))

std_change_df_low <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.01, 0.49, by = 0.01))

z_df_low <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.01, 0.49, by = 0.01))


for (i in 1:49){
  perc <- mean_df_low[i,1] # for measuring amount of precip/negative change (means)
  
  mean_sub <- master_mean%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_s_test <- cor.test(mean_sub$years_until_resettlement,mean_sub$climate_value)
  
  mean_df_low[i,c("cor_coef")] <- m_s_test$estimate
  mean_df_low[i,c("p_value")] <- m_s_test$p.value
  mean_df_low[i,c("site_count")] <- length(unique(mean_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% mean_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% mean_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% mean_sub$ID)
  

  mean_df_low[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  mean_df_low[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  # std precip
  std_sub <- master_std%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_s_test <- cor.test(std_sub$years_until_resettlement,std_sub$climate_value)
  
  std_df_low[i,c("cor_coef")] <- s_s_test$estimate
  std_df_low[i,c("p_value")] <- s_s_test$p.value
  std_df_low[i,c("site_count")] <- length(unique(std_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% std_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% std_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% std_sub$ID)
  
  std_df_low[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  std_df_low[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  
  # mean precip change
  mean_change_sub <- change_mean%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_c_test <- cor.test(mean_change_sub$years_until_resettlement,mean_change_sub$climate_value)
  
  mean_change_df_low[i,c("cor_coef")] <- m_c_test$estimate
  mean_change_df_low[i,c("p_value")] <- m_c_test$p.value
  mean_change_df_low[i,c("site_count")] <- length(unique(mean_change_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% mean_change_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% mean_change_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% mean_change_sub$ID)
  
  mean_change_df_low[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  mean_change_df_low[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  # std precip
  std_change_sub <- change_std%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_c_test <- cor.test(std_change_sub$years_until_resettlement,std_change_sub$climate_value)
  
  std_change_df_low[i,c("cor_coef")] <- s_c_test$estimate
  std_change_df_low[i,c("p_value")] <- s_c_test$p.value
  std_change_df_low[i,c("site_count")] <- length(unique(std_change_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% std_change_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% std_change_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% std_change_sub$ID)
  
  std_change_df_low[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  std_change_df_low[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  
  # z score
  z_score_sub <- z_score%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  z_test <- cor.test( z_score_sub$years_until_resettlement, z_score_sub$climate_value)
  
   z_df_low[i,c("cor_coef")] <- z_test$estimate
   z_df_low[i,c("p_value")] <- z_test$p.value
   z_df_low[i,c("site_count")] <- length(unique(z_score_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% z_score_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% z_score_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% z_score_sub$ID)
  
  z_df_low[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  z_df_low[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
}
```

#### 6.2.2 Examine Correlations for all values at highest percentiles

Similarly, for each climate measure, values at each percentile from ≥ P~51~ to ≥ P~99~ were identified and correlations were also run. These percentiles are especially important for correctly measuring the most extreme experiences of absolute precipitation change.

```{r}
#| echo: false
mean_df_high <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.51, 0.99, by = 0.01))

std_df_high <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.51, 0.99, by = 0.01))

mean_change_df_high <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.51, 0.99, by = 0.01))

std_change_df_high <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.51, 0.99, by = 0.01))

z_df_high <- as.data.frame(matrix(0,49,6))%>%
  setNames(c("percentile","cor_coef","p_value","site_count","longevity","average_size_of_vep_rmblk"))%>%
  mutate(percentile = seq(0.51, 0.99, by = 0.01))

for(i in 1:49){
  perc <- mean_df_high[i,1] # for measuring amount of precip/negative change (means)
  
  # mean precip
  mean_sub <- master_mean%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_s_test <- cor.test(mean_sub$years_until_resettlement,mean_sub$climate_value)
  
  mean_df_high[i,c("cor_coef")] <- m_s_test$estimate
  mean_df_high[i,c("p_value")] <- m_s_test$p.value
  mean_df_high[i,c("site_count")] <- length(unique(mean_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% mean_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% mean_sub$ID)
  
    vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% mean_sub$ID)
  
  mean_df_high[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  mean_df_high[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  # std precip
  std_sub <- master_std%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_s_test <- cor.test(std_sub$years_until_resettlement,std_sub$climate_value)
  
  std_df_high[i,c("cor_coef")] <- s_s_test$estimate
  std_df_high[i,c("p_value")] <- s_s_test$p.value
  std_df_high[i,c("site_count")] <- length(unique(std_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% std_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% std_sub$ID)
  
    vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% std_sub$ID)
  
  std_df_high[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  std_df_high[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  # mean precip change
  mean_change_sub <- change_mean%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_c_test <- cor.test(mean_change_sub$years_until_resettlement,mean_change_sub$climate_value)
  
  mean_change_df_high[i,c("cor_coef")] <- m_c_test$estimate
  mean_change_df_high[i,c("p_value")] <- m_c_test$p.value
  mean_change_df_high[i,c("site_count")] <- length(unique(mean_change_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% mean_change_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% mean_change_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% mean_change_sub$ID)
  
  mean_change_df_high[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  mean_change_df_high[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  # std precip
  std_change_sub <- change_std%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_c_test <- cor.test(std_change_sub$years_until_resettlement,std_change_sub$climate_value)
  
  std_change_df_high[i,c("cor_coef")] <- s_c_test$estimate
  std_change_df_high[i,c("p_value")] <- s_c_test$p.value
  std_change_df_high[i,c("site_count")] <- length(unique(std_change_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% std_change_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% std_change_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% std_change_sub$ID)
  
  std_change_df_high[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  std_change_df_high[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
  
  
  # z score
  z_score_sub <- z_score%>%
    group_by(years_until_resettlement) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  z_test <- cor.test( z_score_sub$years_until_resettlement, z_score_sub$climate_value)
  
   z_df_high[i,c("cor_coef")] <- z_test$estimate
   z_df_high[i,c("p_value")] <- z_test$p.value
   z_df_high[i,c("site_count")] <- length(unique(z_score_sub$ID))
  
  gh_in_sub <- gh%>%
    filter(SITE_NA %in% z_score_sub$ID)
  
  vep_in_sub <- vepii_sites%>%
    filter(UNIQ_ID %in% z_score_sub$ID)
  
  vep_occ_in_sub <- vep_occupation_timeseries_simplified%>%
    filter(ID %in% z_score_sub$ID)
  
  z_df_high[i,c("longevity")] <- mean(c(gh_in_sub$occupation_length,vep_occ_in_sub$occupation_length))
  z_df_high[i,c("average_size_of_vep_rmblk")] <- mean(vep_in_sub$roomblock.area..m2,na.rm=T)
}
```

#### 6.2.1 Plot Results

Combine all of the results, identify correlations with significant p-values, and plot.

```{r}
#| echo: false

mean_df <- rbind(mean_df_low, mean_df_high)%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))
         
std_df<- rbind(std_df_low,std_df_high)%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))

mean_change_df<- rbind(mean_change_df_low,mean_change_df_high)%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))
         
std_change_df<- rbind(std_change_df_low,std_change_df_high)%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))

z_df <- rbind(z_df_low,z_df_high)%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))
  
mean_df_sig <- mean_df%>%
  filter(sig == 1)
std_df_sig <- std_df%>%
  filter(sig == 1)
mean_change_df_sig <- mean_change_df%>%
  filter(sig == 1)
std_change_df_sig <- std_change_df%>%
  filter(sig == 1)
z_df_sig <-z_df%>%
  filter(sig==1)
```

```{r}
#| echo: false

p1 <- ggplot()+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill=NA,color="#4D4D4D80")+
  geom_line(data=mean_df,aes(percentile,cor_coef),col="#780000",lwd=1.2)+
  geom_point(data=mean_df_sig,aes(percentile,cor_coef),col="#780000",size=1.5)+
  
   annotate("text",x=0.48,y=0.85,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=1,lineheight = 0.8)+ 
    annotate("text",x=0.48,y=0.73,label="Low. Precip",size=4, color="#780000",hjust=1)+

  annotate("text",x=0.52,y=-0.7,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=0,lineheight = 0.8)+ 
      annotate("text",x=0.52,y=-0.82,label="High Precip",size=4, color="#780000",hjust=0)+

  labs(x=NULL,
       y="Coorelation Coefficient",
       title= "Precipitation Availability",
       subtitle="A")+
  
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10),
        plot.title=element_text(size=22,hjust=0.5),
        plot.subtitle=element_text(size=17,color="grey30"))

p2 <- ggplot()+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill=NA,color="#4D4D4D80")+
  geom_line(data=z_df,aes(percentile,cor_coef),col="#bb3e03",lwd=1.2)+
  geom_point(data=z_df_sig,aes(percentile,cor_coef),col="#bb3e03",size=1.5)+
  geom_line(data=mean_change_df,aes(percentile,cor_coef),col="#ee9b00",lwd=1.2)+
  geom_point(data=mean_change_df_sig,aes(percentile,cor_coef),col="#ee9b00",size=1.5)+
  
   annotate("text",x=0.48,y=0.85,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=1,lineheight = 0.8)+ 
    annotate("text",x=0.49,y=0.73,label="Negative 5-year Z-scores",size=4, color="#bb3e03",hjust=1)+
   annotate("text",x=0.48,y=0.67,label="Negative Temporal Variability",size=4, color="#ee9b00",hjust=1)+

  annotate("text",x=0.52,y=-0.7,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=0,lineheight = 0.8)+ 
  annotate("text",x=0.52,y=-0.82,label="Positive 5-year Z-scores",size=4, color="#bb3e03",hjust=0)+
  annotate("text",x=0.52,y=-0.88,label="Positive Temporal Variability",size=4, color="#ee9b00",hjust=0)+


  labs(x=NULL,
       y=NULL,
       title= "Directional Precipitation Change",
       subtitle="B")+
  
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(5,10,0,10),
        plot.title=element_text(size=22,hjust=0.5),
        plot.subtitle=element_text(size=17,color="grey30"))

p3<- ggplot()+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill=NA,color="#4D4D4D80")+
  geom_line(data=std_change_df,aes(percentile,cor_coef),col="#005f73",lwd=1.2)+
  geom_point(data=std_change_df_sig,aes(percentile,cor_coef),col="#005f73",size=1.5)+
  geom_line(data=std_df,aes(percentile,cor_coef),col="#94d2bd",lwd=1.2)+
  geom_point(data=std_df_sig,aes(percentile,cor_coef),col="#94d2bd",size=1.5)+

     annotate("text",x=0.48,y=0.85,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=1,lineheight = 0.8)+ 
  annotate("text",x=0.48,y=0.73,label="Less Spatio-temporal Variability",size=4, color="#005f73",hjust=1)+
      annotate("text",x=0.48,y=0.67,label="Less Spatial Variability",size=4, color="#94d2bd",hjust=1)+
  
    annotate("text",x=0.52,y=-0.7,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=0,lineheight = 0.8)+ 
  annotate("text",x=0.52,y=-0.82,label="More Spatio-temporal Variability",size=4, color="#005f73",hjust=0)+
  annotate("text",x=0.52,y=-0.88,label="More Spatial Variability",size=4, color="#94d2bd",hjust=0)+
  
  labs(x=NULL,
       y=NULL,
       title= "Absolute Precipitation Change",
       subtitle="C")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(5,10,0,10),
        plot.title=element_text(size=22,hjust=0.5),
        plot.subtitle=element_text(size=17,color="grey30"))

p4 <- ggplot()+
  geom_point(data=mean_df,aes(percentile, site_count), col="#780000")+
  labs(x=NULL,
       y="No. of Sites",
       subtitle= "D")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(0,7000),breaks=c(1000,3000,5000,7000))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17))

p5 <- ggplot()+
  geom_point(data=mean_change_df,aes(percentile, site_count),col="#ee9b00")+
  geom_point(data=z_df,aes(percentile, site_count),col="#bb3e03")+
  labs(x=NULL,
       y=NULL,
       subtitle= "E")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(0,7000),breaks=c(1000,3000,5000,7000))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

p6<-ggplot()+
  geom_point(data=std_df,aes(percentile, site_count),col="#005f73")+
  geom_point(data=std_change_df,aes(percentile, site_count),col="#94d2bd")+
  labs(x=NULL,
       y=NULL,
       subtitle= "F")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(0,7000),breaks=c(1000,3000,5000,7000))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))


p7 <- ggplot()+
  geom_point(data=mean_df,aes(percentile, longevity), col="#780000")+
  labs(x=NULL,
       y="Average Length of \nOccupation (years)",
       subtitle= "G")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(100,305),breaks=c(100,200,300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

p8<-ggplot()+
  geom_point(data=mean_change_df,aes(percentile,longevity),col="#ee9b00")+
  geom_point(data=z_df,aes(percentile,longevity),col="#bb3e03")+
  labs(x=NULL,
       y=NULL,
       subtitle= "H")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(100,305),breaks=c(100,200,300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

p9<-ggplot()+
  geom_point(data=std_df,aes(percentile,longevity),col="#005f73")+
  geom_point(data=std_change_df,aes(percentile,longevity),col="#94d2bd")+
  labs(x=NULL,
       y=NULL,
       subtitle= "I")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(100,305),breaks=c(100,200,300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title = element_text(size=15,margin=margin(t=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))


p10 <- ggplot()+
  geom_point(data=mean_df,aes(percentile, average_size_of_vep_rmblk), col="#780000")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(250,450),breaks=c(250,350,450))+
  labs(x="Percentile",
       y="Average Roomblock Size \n(sq. m)",
       subtitle= "J")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        axis.title.x = element_text(size=15,margin=margin(t=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

p11 <- ggplot()+
  geom_point(data=mean_change_df,aes(percentile,average_size_of_vep_rmblk),col="#ee9b00")+
  geom_point(data=z_df,aes(percentile,average_size_of_vep_rmblk),col="#bb3e03")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(250,450),breaks=c(250,350,450))+
  labs(x="Percentile",
       y=NULL,
       subtitle= "K")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.x = element_text(size=15,margin=margin(t=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

p12 <- ggplot()+
  geom_point(data=std_df,aes(percentile,average_size_of_vep_rmblk),col="#005f73")+
  geom_point(data=std_change_df,aes(percentile,average_size_of_vep_rmblk),col="#94d2bd")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(250,450),breaks=c(250,350,450))+
  labs(x="Percentile",
       y=NULL,
       subtitle= "L")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.x = element_text(size=15,margin=margin(t=5)),
        plot.margin = margin(5,10,0,10),
        plot.subtitle=element_text(size=17,color="grey30"))

png('./FIGURES/Figure_4.png',height=20000,width=24000,res=1200)
ggarrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,
          ncol=3,nrow=4,align="hv",heights=c(1.65,0.7,0.7,0.7))
dev.off()

```

Export.

```{r}
#| echo: false

#OPTIONAL: 
write.csv(mean_df,"./OUTPUT/precip_availability_cor.csv",row.names=F)
write.csv(std_df,"./OUTPUT/precip_spatialvariability_cor.csv",row.names=F)
write.csv(mean_change_df,"./OUTPUT/precip_temporalvariability_cor.csv",row.names=F) 
write.csv(std_change_df,"./OUTPUT/precip_spatiotemporalvariability_cor.csv",row.names=F)
write.csv(z_df,"./OUTPUT/precip_5yrzscore_cor.csv",row.names=F)
```

## 7.0 Analyze Dendrochronological Data

Dendrochronological data was accessed from the Digital Archaeological Record (tDAR) and can be accessed [here](https://doi.org/10.6067/XCV82J6D7B):

1.  Compiled Tree-ring Dates from the Southwestern United States (Restricted). Timothy A. Kohler, R. Kyle Bocinsky. Science Advances. 2016 ( tDAR id: 399314) ; [doi:10.6067/XCV82J6D7B](https://doi.org/10.6067/XCV82J6D7B)

These data include 32,863 tree ring dates from archaeological sites in New Mexico, Arizon, Colorado and Utah. All date determination in the data set were made by the Laboratory of Tree Ring Research. Nearly all of these data (n =29,311 dates) were used by Bocinsky et al. (2016, [DOI](https://doi.org/10.1126/sciadv.1501532)) for PaeloCAR. For this study, only dates with a reliable or cutting or near cutting date were used, limiting the sample to 7,449 tree ring dates.

### 7.1 Import data

```{r}
#| echo: false
dendro_dates <- read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/SENSITIVE SITE LOCATIONS/BOCINSKY_ET_AL_2015_TREE_RINGS.csv")%>%
  st_as_sf(.,coords=c("LON","LAT"),crs=4326)%>%
  st_intersection(.,study_extent)%>%
  subset(.,Outer_Symbol == "V"|Outer_Symbol=="RLB"|Outer_Symbol=="R"|Outer_Symbol=="+B"|Outer_Symbol=="++G"|Outer_Symbol=="RG"|Outer_Symbol=="+V"|Outer_Symbol=="L"|Outer_Symbol=="+RLB"|Outer_Symbol=="CB"|Outer_Symbol=="+L"|Outer_Symbol=="+LB"|Outer_Symbol=="LGB"|Outer_Symbol=="C"|Outer_Symbol=="CL"|Outer_Symbol=="+R"|Outer_Symbol=="+CL"|Outer_Symbol=="CGB"|Outer_Symbol=="RLG"|Outer_Symbol=="GB"|Outer_Symbol=="VL"|Outer_Symbol=="GLB"|Outer_Symbol=="+GLB"|Outer_Symbol=="BB"|Outer_Symbol=="+LG"|Outer_Symbol=="+RGB"|Outer_Symbol=="CLB"|Outer_Symbol=="c"|Outer_Symbol=="r"|Outer_Symbol=="cB"|Outer_Symbol=="+r"|Outer_Symbol=="rB"|Outer_Symbol=="+VL"|Outer_Symbol=="CG"|Outer_Symbol=="+CG"|Outer_Symbol=="RC"|Outer_Symbol=="LB "|Outer_Symbol=="rGB"|Outer_Symbol== "RB COM"|Outer_Symbol== "R COMP"|Outer_Symbol== "R INC")%>%
  subset(.,!is.na(Site_Number))%>%
  mutate(ID=1:nrow(.))
```

### 7.2 Synthesize Climate Data

Unique site locations were identified and climate conditions at each unique site location were calculated following the same steps as above (Step 4).

```{r}
#| echo: false

duplicates <- st_coordinates(dendro_dates)%>%
  duplicated()

dendro_dates_unique_loc <- dendro_dates[!duplicates,]
```

```{r}
#| echo: false

for(i in 1:nrow(dendro_dates_unique_loc)){
  site <- dendro_dates_unique_loc[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- dendro_dates_unique_loc[i,c("ID")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("mean_precipitation","year"))
  
  spatial_variability <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=600:1351)%>%
    setNames(c("spatial_variability","year"))
  
  temporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("temporal_variability","year"))

  spatiotemporal_variability <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=601:1351)%>%
    setNames(c("spatiotemporal_variability","year"))
  
    z_scores <- mean_precipitation%>%
    mutate(five_year_av = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=mean,fill=NA),
            five_year_sd = rollapplyr(mean_precipitation,width=list(-(1:5)),FUN=sd,fill=NA),precip_5yr_zscore = (mean_precipitation-five_year_av)/five_year_sd)%>%
  select(c("year","precip_5yr_zscore"))

  summary <- merge(mean_precipitation,spatial_variability,by="year")%>%
    merge(.,temporal_variability,by="year")%>%
    merge(.,spatiotemporal_variability,by="year")%>%
    merge(.,z_scores,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_dendro <- summary
  }else {
    summary_all_dendro <- merge(summary_all_dendro,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   spatial_variability,
   temporal_variability,
   spatiotemporal_variability)

# OPTIONAL:
write.csv(summary_all_dendro,"./OUTPUT/climatevariability_dendro.csv",row.names=F)
```

### 7.3 Synthesize occupation timeline

For this analysis, tree ring dates with a cutting event in each year were summed. Fifteen years were missing any tree ring dates. For this time series to be easily subjected to a Granger causality test, the number of tree ring dates in missing years were linearly interpolated from the number of samples for the year before and the year after.

```{r}
#| echo: false
dendro_df <-  st_drop_geometry(dendro_dates)%>%
  select(c("ID","Outer_Date_AD"))%>%
  setNames(c("ID","year"))%>%
  mutate(count= 1)%>%
  group_by(year)%>%
  reframe(count=sum(count))%>%
  filter(., year >= 1033 & year <= 1335)%>%
  complete(year = 1033:1335)%>%
  mutate(interpolated = ifelse(is.na(count),1,0),
         count_raw = count,
         count = zoo::na.approx(count))

# OPTIONAL
write.csv(dendro_df,"./OUTPUT/dendro_activity_timeseries.csv")
```

### 7.4 Combine Climate and Occupation

The number of samples for many years only include tree rings from 1-2 sites. Therefore, the mean for each climate measure in each year was calculated (rather than a percentile cut off of, say, ≤ P~1~ ).

```{r}
#| echo: false
summary_dendro <- summary_all_dendro%>%
  filter(., year >= 1033 & year <= 1335)%>%
  gather(key=ID,value=climate_value,-c("year","variable"))%>%
  mutate(ID=as.numeric(sub("^X","",ID)))


dendro_mean <- subset(summary_dendro, variable == "mean_precipitation")%>%
    group_by(year) %>%
    summarise(mean_precipitation=mean(climate_value,na.rm=T))
    

dendro_std <- subset(summary_dendro, variable == "spatial_variability")%>%
    group_by(year) %>%
    summarise(spatial_variability=mean(climate_value,na.rm=T))


dendro_change_mean <- subset(summary_dendro, variable == "temporal_variability")%>%
    group_by(year) %>%
    summarise(temporal_variability=mean(climate_value,na.rm=T))


dendro_change_std <- subset(summary_dendro, variable == "spatiotemporal_variability")%>%
    group_by(year) %>%
    summarise(spatiotemporal_variability=mean(climate_value,na.rm=T))


dendro_z <- subset(summary_dendro, variable == "precip_5yr_zscore")%>%
    group_by(year) %>%
    summarise(precip_5yr_zscore=mean(climate_value,na.rm=T))

dendro_combined_timeseries <- merge(dendro_df,dendro_mean,by=c("year"))%>%
merge(.,dendro_std,by=c("year"))%>%
merge(.,dendro_change_mean,by=c("year"))%>%
merge(.,dendro_change_std,by=c("year"))%>%
merge(.,dendro_z,by=c("year"))
```

### 7.5 Granger Causality Test

Granger causality tests are conducted to compare the predictability between all five climate measures and future instances of building activity. First, the complete tree ring and climate time series were submitted to an Augmented Dickey-Fuller test, because Granger causality tests are intended for stationary series. Each time series was shown to be stationary. NOTE: Because these series are several hundred years long, they are very likely to be stationary; however, this might mean key instances of change where patterns skew rapidly, are under appreciated.

#### 7.5.1 Augmented Dickey-Fuller tests

```{r}
# p < 0.05 = stationary so good for granger causality
adf.test(dendro_combined_timeseries[,2])
adf.test(dendro_combined_timeseries[,5])
adf.test(dendro_combined_timeseries[,6])
adf.test(dendro_combined_timeseries[,7])
adf.test(dendro_combined_timeseries[,8])
adf.test(dendro_combined_timeseries[,9])
```

#### 7.5.2 Loop of Granger Causality tests

This loop compares each climate measure against a 1-25 year lag in tree-ring building activity. Granger causality p-values are recorded for each test and recorded in the object `gc_results`.

```{r}
#| echo: false

gc_results <- data.frame(year_lag=1:25,
                         mean_precipitation=NA,
                         temporal_variability=NA,
                         spatial_variability=NA,
                         spatiotemporal_variability=NA,
                         z_score=NA)

for(i in c("mean_precipitation","spatial_variability","temporal_variability","spatiotemporal_variability","precip_5yr_zscore")){
  
  dendro_sample <- dendro_combined_timeseries%>%
    select(all_of(c("count",i)))
  
  for(j in 1:25){
      gt<-grangertest(dendro_sample[,1] ~ dendro_sample[,2], order = j)
      gc_results[j,c(i)] <- gt$`Pr(>F)`[2]
  }
}
```

Plot the results.

```{r}
#| echo: false

p1<- ggplot()+
  geom_line(data=dendro_df,aes(year,count,group=1,color=as.factor(interpolated)),linewidth=1)+
  scale_color_manual(values=c("#8AB17D","#EE8959"))+
  labs(title="Number of Tree Ring cutting Dates by Year",
       x=NULL,
       y="Number of Cutting Dates",
       subtitle="orange line represents interpolated values",
  )+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        plot.title = element_text(size=17,margin=margin(t=10,b=3)),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,5,10),
        legend.position = "none")

p2<-ggplot()+
  geom_line(data=dendro_mean,aes(year,mean_precipitation),col="#bb3e03")+
  labs(x=NULL,
       y="Annual Mean Precipitation \n(mm)",
       title=NULL)+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10))

p3<-ggplot()+
  geom_line(data=dendro_change_mean,aes(year,temporal_variability),col="#ee9b00")+
  labs(x=NULL,
       y="Temporal Variability \n(mm)",
       title=NULL)+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10))


p4<-ggplot()+
  geom_line(data=dendro_z,aes(year,precip_5yr_zscore),col="#780000")+
  labs(x=NULL,
       y="5-year Z-score",
       title=NULL)+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10))

p5<-ggplot()+
  geom_line(data=dendro_std,aes(year,spatial_variability),col="#94d2bd")+
  labs(x=NULL,
       y="Temporal Variability \n(standard deviation)",
       title=NULL)+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10))

p6<-ggplot()+
  geom_line(data=dendro_change_std,aes(year,spatiotemporal_variability),col="#005f73")+
  labs(x="Year (CE)",
       y="Spatio-temporal Variability \n(standard deviation)",
       title=NULL)+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.x = element_text(size=15,margin=margin(t=5)),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        plot.margin = margin(5,10,0,10))


p7<-ggplot()+
  geom_line(data=gc_results,aes(year_lag,precip_5yr_zscore),col="#780000")+
  geom_point(data=gc_results,aes(year_lag,precip_5yr_zscore),col="#780000")+
  geom_line(data=gc_results,aes(year_lag,mean_precipitation),col="#bb3e03")+
  geom_point(data=gc_results,aes(year_lag,mean_precipitation),col="#bb3e03")+
  geom_line(data=gc_results,aes(year_lag,spatial_variability),col="#94d2bd")+
  geom_point(data=gc_results,aes(year_lag,spatial_variability),col="#94d2bd")+
  geom_line(data=gc_results,aes(year_lag,temporal_variability),col="#ee9b00")+
  geom_point(data=gc_results,aes(year_lag,temporal_variability),col="#ee9b00")+
  geom_line(data=gc_results,aes(year_lag,spatiotemporal_variability),col="#005f73")+
  geom_point(data=gc_results,aes(year_lag,spatiotemporal_variability),col="#005f73")+
  geom_hline(yintercept=0.05,col="grey30",linetype="dotted")+
  labs(x="Year Lag",
       y="Granger Causality \nP-value",
       title=NULL)+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=5)),
        axis.title.x = element_text(size=15,margin=margin(t=5)),
        plot.margin = margin(5,10,5,10))


png('./FIGURES/Figure_5.png',height=26000,width=19000,res=1200)
ggarrange(p1,p2,p3,p4,p5,p6,p7,
          ncol=1,nrow=7,align="hv",heights=c(1.25,1,1,1,1,1,1.75))
dev.off()
```

## A. Supplemental Figures

Below is a supplemental GIF of all climate variability measures plotted for the study period 600-1350.

```{r}
#| echo: false  

df <- as.data.frame(r_var,xy=T)%>%
  pivot_longer(cols = -c(x, y), names_to = "layer",values_to = "value")%>%
  mutate(layer_num = as.numeric(gsub("PPT_", "", layer)))  

p <- ggplot() +   
  geom_raster(data=df, aes(x = x, y = y, fill = value)) +   
  scale_fill_gradientn(colors=c("#bb3e03","#ca6702","#ee9b00","#ffe6a750","#94d2bd","#0a9396","#005f73"), rescaler = ~ scales::rescale_mid(.x, mid = 0), name = "Precipitation  Variability (mm)")+   
  geom_sf(data=study_extent,col="grey20",fill=NA)+   
  geom_sf(data=four_corners,col="grey70",fill=NA,linetype="dotted")+   
  coord_equal() +   
  labs(title = "Year (CE): {frame_time}", 
       x=NULL,
       y=NULL) +   
  transition_time(layer_num) +   
  ease_aes("linear")+   
  theme_minimal()  

animate(p, fps = 5,nframes = 701, width = 750, height = 600)  anim_save("./FIGURES/raster_precip_variability_600to1350.gif")  
```
