---
title: "Field and DePlata-Peterson 2026"
format: html
editor: visual
---

# Research compendium for Field and DePlata-Peterson (2026)

This R-based research compendium supports the...

View the raw source code and output at: provide link

**When using the code included in this research compendium, please cite *all* of the following:**

> Field, S and M. DePlata-Peterson. 2026. *Article Name*. Journal Location, In review.
>
> Field, S and M. DePlata-Peterson. 2026. *Article Name*, Version 1.0.0. Zenodo. \[DOI\]

## Step 1: Set Up Environment

```{r setup, include=FALSE}
#| echo: false
knitr::opts_knit$set(root.dir = "C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/PROJECT/CLIMATE_VARIABILITY_REDUX")
```

```{r}
#| echo: false
packages <-c('terra','sf','ggplot2','tidyterra','dplyr','tidyr','gifski','gganimate')

for(p in packages) if(p %in% rownames(installed.packages()) == F) { install.packages(p) }
for(p in packages) suppressPackageStartupMessages(library(p,quietly=T,character.only=T))

theme_set(theme_bw())
```

### Establish Unique Functions

This function is designed to identify an occupation end date and build a time series counting down until disoccupation.

```{r}
#| echo: false
find_years_until_disoccupation <- function(x) {
  n <- length(x)
  out <- integer(n)
  counter <- 0
  for (i in seq(n, 1)) {
    if (x[i] == 0) {
      counter <- 0
    } else {
      counter <- counter + 1
    }
    out[i] <- counter
  }
  out
}
```

## Step 2: Study Area and Paleoclimate Raster

Study area derived as extent of sites used in study.

Variability raster downloaded from....

```{r}
#| echo: false
study_extent <- st_read("./DATA/study_extent.shp")%>%
  st_transform(.,crs=4269)

four_corners <- st_read("C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/DATA_GENERAL/SPATIAL/NORTH_AMERICA_REGIONS/SOUTHWEST/STATE_BOUNDARIES/four_corners_states.shp")%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)

```

```{r}
#| echo: false
# r <- terra::rast("./DATA/CLIMATE/PPT.tif")
# 
# # calculate year-to-year variability and export
# r_var <- (subset(r, 1:(nlyr(r)-1)) - subset(r, 2:nlyr(r))) * -1
# 
# writeRaster(r_var,"./DATA/CLIMATE/PPT_annualvariability.tif")
```

After variability assessment is computed/exported, read in directly.

```{r}
#| echo: false
r_var <- terra::rast("./DATA/CLIMATE/PPT_annualvariability.tif")
```

### Plotting Variability Raster

```{r}
#| echo: false

r_var1 <- r_var[[600:1300]]

df <- as.data.frame(r_var1,xy=T)%>%
  pivot_longer(cols = -c(x, y), names_to = "layer",values_to = "value")%>%
  mutate(layer_num = as.numeric(gsub("PPT_", "", layer)))

p <- ggplot() +
  geom_raster(data=df, aes(x = x, y = y, fill = value)) +
  scale_fill_gradientn(colors=c("#bb3e03","#ca6702","#ee9b00","#ffe6a780","#94d2bd","#0a9396","#005f73"), rescaler = ~ scales::rescale_mid(.x, mid = 0),
name = "Precipitation 
Variability (mm)")+
  geom_sf(data=study_extent,col="grey20",fill=NA)+
  geom_sf(data=four_corners,col="grey70",fill=NA,linetype="dotted")+
  #coord_equal() +
  labs(title = "Year (CE): {frame_time}",
       x=NULL,
       y=NULL) +
  transition_time(layer_num) +
  ease_aes("linear")+
  theme_minimal()

animate(p, fps = 25,nframes = 701, width = 750, height = 600)

anim_save("./FIGURES/raster_precip_variability_600to1300.gif")

rm(r_var1)
```

## Step 3: Chaco Great Houses

Accessed from:

### Import data

```{r}
#| echo: false

chaco_greathouses <- st_read("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/VanDykeEtAl2015_WGS84.shp")%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)

outlier_database <-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/OutlierAttributes_Public_CRA.csv")%>%
  select("Site.name","Site.number.","Est..great.house.start.date.","Est..great.house.end.date.")%>%
  setNames(c("SITE_NA","SITE_NUMBE","BEGIN_DATE","END_DATE"))

gh <- merge(chaco_greathouses,outlier_database,by="SITE_NA")%>%
  filter(!is.na(BEGIN_DATE))%>%
  filter(!is.na(END_DATE))%>%
  subset(BEGIN_DATE != "No Data")%>%
  subset(END_DATE != "No Data")%>%
  select(-c("SITE_NUMBE.x","SITE_NUM_1","PROJECTED_","PROJECTE_1","PROJECTE_2"))%>%
  mutate(END_DATE = as.numeric(END_DATE),
         BEGIN_DATE = as.numeric(BEGIN_DATE),
         occupation_length = END_DATE - BEGIN_DATE)%>%
  subset(occupation_length >= 1)

rm(chaco_greathouses,outlier_database)
```

### Synthesize climate data

**NOTE:** in most places all of the change is either negative or positive, so usually absolute change is the additive inverse. However, in some places where some cells change positively and some negatively, the absolute change is different.

```{r}
# for(i in 1:nrow(gh)){
#   site <- gh[i,]%>%
#     st_buffer(.,2000)%>%
#     st_transform(.,crs=4326)
#   
#   site_name <- gh[i,c("SITE_NA")]%>%
#                 st_drop_geometry()%>%
#                 as.character()
#   
#   site_var <- terra::extract(r_var,site,exact=T)
#   
#   
#   mean_change <- site_var%>%
#     select(-c("ID","fraction"))%>%
#     summarise_all(mean,na.rm=T)%>%
#     t()%>%
#     as.data.frame()%>%
#     mutate(year=2:2000)%>%
#     setNames(c("precip_mean_change","year"))
# 
#   std_change <- site_var%>%
#     select(-c("ID","fraction"))%>%
#     summarise_all(sd,na.rm=T)%>%
#     t()%>%
#     as.data.frame()%>%
#     mutate(year=2:2000)%>%
#     setNames(c("precip_std_change","year"))
# 
#   absolute_change <- abs(site_var)%>%
#     select(-c("ID","fraction"))%>%
#     summarise_all(mean,na.rm=T)%>%
#     t()%>%
#     as.data.frame()%>%
#     mutate(year=2:2000)%>%
#     setNames(c("precip_absolutemean_change","year"))
#   
# 
#   summary <- merge(mean_change,std_change,by="year")%>%
#     merge(.,absolute_change,by="year")%>%
#     gather(.,key="variable",value="value",-year)%>%
#     setNames(c("year","variable",site_name))
# 
#   if (i == 1){
#     summary_all_greathouses <- summary
#   }else {
#     summary_all_greathouses <- merge(summary_all_greathouses,summary, by=c("year","variable"))
#   }
# }
# 
# write.csv(summary_all_greathouses,"./OUTPUT/climatevariability_greathouses.csv")
```

## Step 4: VEP II North Study Area Site Data

Accessed from:

### Import data

```{r}
#| echo: false
# example script
vepii_sites<-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/db_vepii_n.csv")%>%
  st_as_sf(.,coords=c("UTMEast","UTMNorth"),crs=26912)%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)%>%
  mutate(UNIQ_ID = 1:nrow(.))

vep_df <- read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/db_vepii_n.csv")%>%
  select(c("ph6","ph7","ph8","ph9","ph10","ph11","ph12",
           "ph13","ph14","ph15","ph16","ph17","ph18","ph19"))%>%
  mutate(UNIQ_ID = 1:nrow(.))%>%
  gather(phase,value,-UNIQ_ID)%>%
  mutate(value = ifelse(value<0,0,value))

vep_dates <- read.csv("./DATA/VEP_dates.csv")

vep <- left_join(vep_df,vep_dates,c("phase"="PHASE"))

rm(vep_dates)

vep<- arrange(vep,desc(UNIQ_ID),BEGIN_D)

expanded_vep <- vep %>%
  rowwise() %>%
  mutate(year = list(BEGIN_D:END_DAT)) %>%
  unnest(year) %>%
  select(UNIQ_ID, year, value) 

vep_years <- expanded_vep%>%
  group_by(UNIQ_ID)%>%
  mutate(year_until = find_years_until_disoccupation(value))%>%
  setNames(c("UNIQ_ID","Year","occupation","years_until_disoccupation"))

```

### Synthesize climate data

```{r}
#| echo: false
# example script 

```

## Step 5: Synthesize Chaco and VEP data

```{r}
#| echo: false
```
