---
title: "Field and DePlata-Peterson 2026"
format: html
editor: visual
---

# Research compendium for Field and Dean (2026)

This R-based research compendium supports the...

View the raw source code and output at: provide link

**When using the code included in this research compendium, please cite *all* of the following:**

> Field, S and M. DePlata-Peterson. 2026. *Article Name*. Journal Location, In review.
>
> Field, S and M. DePlata-Peterson. 2026. *Article Name*, Version 1.0.0. Zenodo. \[DOI\]

## Step 1: Set Up Environment

```{r}
#| echo: false
packages <-c('terra','sf','ggplot2','tidyterra','dplyr','tidyr','gifski','gganimate')

for(p in packages) if(p %in% rownames(installed.packages()) == F) { install.packages(p) }
for(p in packages) suppressPackageStartupMessages(library(p,quietly=T,character.only=T))

theme_set(theme_bw())
```

```{r}
#| echo: false
setwd("C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/PROJECT/CLIMATE_VARIABILITY_REDUX")
```

## Step 2: Study Area and Paleoclimate Raster

Study area derived as extent of sites used in study.

Variability raster downloaded from....

```{r}
#| echo: false
study_extent <- st_read("./DATA/study_extent.shp")%>%
  st_transform(.,crs=4269)

r <- terra::rast("./DATA/CLIMATE/PPT.tif")

# calculate year-to-year variability and export
r_var <- (subset(r, 1:(nlyr(r)-1)) - subset(r, 2:nlyr(r))) * -1

writeRaster(r_var,"./DATA/CLIMATE/PPT_annualvariability.tif")
```

After variability assessment is computed/exported, read in directly.

```{r}
#| echo: false
r_var <- terra::rast("./DATA/CLIMATE/PPT_annualvariability.tif")
```

### Plotting Variability Raster

```{r}
#| echo: false

r_var1 <- r_var[[600:1300]]

df <- as.data.frame(r_var1,xy=T)%>%
  pivot_longer(cols = -c(x, y), names_to = "layer",values_to = "value")%>%
  mutate(layer_num = as.numeric(gsub("PPT_", "", layer)))

p <- ggplot(df, aes(x = x, y = y, fill = value)) +
  geom_raster() +
  scale_fill_gradientn(colors=c("#bb3e03","#ca6702","#ee9b00","#FEFAE040","#94d2bd","#0a9396","#005f73"), rescaler = ~ scales::rescale_mid(.x, mid = 0),
name = "Precipitation 
Variability (mm)")+
  #coord_equal() +
  labs(title = "Year (CE): {frame_time}",
       x=NULL,
       y=NULL) +
  transition_time(layer_num) +
  ease_aes("linear")+
  theme_minimal()

animate(p, fps = 1,nframes = 7, width = 750, height = 600)

anim_save("./FIGURES/raster_precip_variability_600to1300.gif")

```

## Step 3: Chaco Great Houses

Accessed from:

```{r}
#| echo: false
# example script

```

## Step 4: VEP II North Study Area Site Data

Accessed from:

```{r}
#| echo: false
# example script

```
