---
title: "Field and DePlata-Peterson 2026"
format: html
editor: visual
---

# Research compendium for Field and DePlata-Peterson (2026)

This R-based research compendium supports the...

View the raw source code and output at: provide link

**When using the code included in this research compendium, please cite *all* of the following:**

> Field, S and M. DePlata-Peterson. 2026. *Variability as a Driver of Re-settlement in the northern US Southwest*. Journal of Archaeological Science, In Prep.
>
> Field, S and M. DePlata-Peterson. 2026. *Research Compendium for Variability as a Driver of Re-settlement in the northern US Southwest*., Version 1.0.0. Zenodo. \[DOI\]

## 1.0 Set Up Environment

```{r setup, include=FALSE}
#| echo: false
knitr::opts_knit$set(root.dir = "C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/PROJECT/CLIMATE_VARIABILITY_REDUX")
```

```{r}
#| echo: false
packages <-c('terra','sf','ggplot2','tidyterra','dplyr','tidyr','gifski','gganimate','ggforce','ggpubr','scales','lmtest','tseries')

for(p in packages) if(p %in% rownames(installed.packages()) == F) { install.packages(p) }
for(p in packages) suppressPackageStartupMessages(library(p,quietly=T,character.only=T))

theme_set(theme_bw())
```

### 1.1 Establish Unique Functions

1.  `find_years_until_disoccupation` - this function is designed to identify an occupation end date and build a time series counting down until disoccupation. b

```{r}
#| echo: false
find_years_until_disoccupation <- function(x) {
  n <- length(x)
  out <- integer(n)
  counter <- 0
  for (i in seq(n, 1)) {
    if (x[i] == 0) {
      counter <- 0
    } else {
      counter <- counter + 1
    }
    out[i] <- counter
  }
  out
}
```

## 2.0 Study Area and Paleoclimate Raster

Study area derived as extent of sites used in study.

Variability raster downloaded from....

```{r}
#| echo: false
study_extent <- st_read("./DATA/study_extent.shp")%>%
  st_transform(.,crs=4269)

four_corners <- st_read("C:/Users/sfield3/OneDrive - University of Wyoming/FIELD LAB HOME/DATA_GENERAL/SPATIAL/NORTH_AMERICA_REGIONS/SOUTHWEST/STATE_BOUNDARIES/four_corners_states.shp")%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)

```

```{r}
#| echo: false
# r <- terra::rast("./DATA/CLIMATE/PPT.tif")
# 
# # calculate year-to-year variability and export
# r_var <- (subset(r, 1:(nlyr(r)-1)) - subset(r, 2:nlyr(r))) * -1
# 
# writeRaster(r_var,"./DATA/CLIMATE/PPT_annualvariability.tif")
```

After variability assessment is computed/exported, read in directly.

```{r}
#| echo: false
r <- terra::rast("./DATA/CLIMATE/PPT.tif")
r_var <- terra::rast("./DATA/CLIMATE/PPT_annualvariability.tif")
```

### 2.1 Plotting Variability Raster

```{r}
#| echo: false

r_var1 <- r_var[[600:1300]]

df <- as.data.frame(r_var1,xy=T)%>%
  pivot_longer(cols = -c(x, y), names_to = "layer",values_to = "value")%>%
  mutate(layer_num = as.numeric(gsub("PPT_", "", layer)))

p <- ggplot() +
  geom_raster(data=df, aes(x = x, y = y, fill = value)) +
  scale_fill_gradientn(colors=c("#bb3e03","#ca6702","#ee9b00","#ffe6a750","#94d2bd","#0a9396","#005f73"), rescaler = ~ scales::rescale_mid(.x, mid = 0),
name = "Precipitation 
Variability (mm)")+
  geom_sf(data=study_extent,col="grey20",fill=NA)+
  geom_sf(data=four_corners,col="grey70",fill=NA,linetype="dotted")+
  #coord_equal() +
  labs(title = "Year (CE): {frame_time}",
       x=NULL,
       y=NULL) +
  transition_time(layer_num) +
  ease_aes("linear")+
  theme_minimal()

animate(p, fps = 5,nframes = 701, width = 750, height = 600)

anim_save("./FIGURES/raster_precip_variability_600to1300.gif")

rm(r_var1)
```

## 3.0 Chaco Great Houses

Accessed from Van Dyke and Heitman (2016) and Chaco Research Archive:

1.  `chaco_greathouses`: this is the geodatabase of Chaco greathouses, collected by... **Note:** These data are not publicly available.
2.  `outlier_database`: this is the Chacoan Outlier database, synthesized by Kantner et al. at the "Chaco World Conference" in 1999. The database was maintained by Kantner (2001-2007), receiving contributions from Paul Reed, R.G. Matson, Bill Lipe, Mark Varien, Scott Ortman, and Susan Ryan. The non-spatial data attributes are publicly accessible at the Chaco Research Archive. Click [here](https://www.chacoarchive.org/cra/outlier-database/) to access.

### 3.1 Import data

```{r}
#| echo: false

chaco_greathouses <- st_read("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/VanDykeEtAl2015_WGS84.shp")%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)

outlier_database <-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/OutlierAttributes_Public_CRA.csv")%>%
  select("Site.name","Site.number.","Est..great.house.start.date.","Est..great.house.end.date.")%>%
  setNames(c("SITE_NA","SITE_NUMBE","BEGIN_DATE","END_DATE"))

gh <- merge(chaco_greathouses,outlier_database,by="SITE_NA")%>%
  subset(SITE_T == "GH_OUTLIER") %>%
  filter(!is.na(BEGIN_DATE))%>%
  filter(!is.na(END_DATE))%>%
  subset(BEGIN_DATE != "No Data")%>%
  subset(END_DATE != "No Data")%>%
  select(-c("SITE_NUMBE.x","SITE_NUM_1","PROJECTED_","PROJECTE_1","PROJECTE_2"))%>%
  mutate(END_DATE = as.numeric(END_DATE),
         BEGIN_DATE = as.numeric(BEGIN_DATE),
         occupation_length = END_DATE - BEGIN_DATE)%>%
  subset(occupation_length >= 1)

rm(chaco_greathouses,outlier_database)
```

### 3.2 Synthesize climate data

**NOTE:** in most places all of the change is either negative or positive, so usually absolute change is the additive inverse. However, in some places where some cells change positively and some negatively, the absolute change is different.

#### 3.2.1 Measures of Climate Stress

This analysis is used to tabulate several measures:

1.  `mean_precipitation` - this is the mean annual precipitation within each catchment.
    1.  This measure is used to test the assumption that years with lower mean precipitation (e.g., the most arid years) are good predictors of social stress.
2.  `std_precipitation` - this is the annual standard deviation within each catchment.
    1.  This measure is used to test the assumption that years with more variable precipitation (e.g., years when catchment is most varied) are good predictors of social stress.
3.  `mean_precipitation_change` - this is the mean year-to-year precipitation variability within each catchment.
    1.  This measure is used to test the assumption that instances of greater negative year-to-year change (e.g., the most unexpectedly arid years) are good predictors of social stress.
4.  `std_precipitation_change` - this is the standard deviation year-to-year precipitation variability within each catchment.
    1.  This measure is used to test the assumption that instances of greater change (regardless of direction) are good predictors of social stress.

```{r}
for(i in 1:nrow(gh)){
  site <- gh[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- gh[i,c("SITE_NA")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("mean_precipitation","year"))
  
  std_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("std_precipitation","year"))
  
  mean_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("mean_precipitation_change","year"))

  std_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("std_precipitation_change","year"))

  
  summary <- merge(mean_precipitation,std_precipitation,by="year")%>%
    merge(.,mean_precipitation_change,by="year")%>%
    merge(.,std_precipitation_change,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_greathouses <- summary
  }else {
    summary_all_greathouses <- merge(summary_all_greathouses,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   std_precipitation,
   mean_precipitation_change,
   std_precipitation_change)

write.csv(summary_all_greathouses,"./OUTPUT/climatevariability_greathouses.csv")
```

### 3.3 Synthesize occupation timeline

Update occupation timeline to know years until disoccupation for all great houses.

```{r}
#| echo: false
gh_occupation_timeseries <- gh %>%
  rowwise() %>%
  st_drop_geometry()%>%
  mutate(year = list(BEGIN_DATE:END_DATE)) %>%
  unnest(year) %>%
  select(c("SITE_NA", "year"))%>%
  group_by(SITE_NA)%>%
  mutate(year_until = find_years_until_disoccupation(year)-1)%>%
  setNames(c("ID","year","years_until_disoccupation"))

write.csv(gh_occupation_timeseries,"./OUTPUT/greathouses_occupation_timeseries.csv")

rm(gh)
```

### 3.4 Compare occupation and climate variables

Results showing that std and change better predictors than how much.

```{r}
#| echo: false

summary_all_greathouses <- read.csv("./OUTPUT/climatevariability_greathouses.csv")%>%select(-c("X"))
gh_occupation_timeseries <- read.csv("./OUTPUT/greathouses_occupation_timeseries.csv")%>%select(-c("X"))

gh_climate_timeseries<-summary_all_greathouses%>%
  gather(key=ID,value=climate_value,-c("year","variable"))

gh_combined_timeseries <- merge(gh_climate_timeseries,gh_occupation_timeseries,by=c("ID","year"))

# mutate(years_until_dis_bin = cut(years_until_disoccupation, breaks = c(-Inf,20,40,60,80,100,120,140,160,180,200,Inf),labels=c("0-20","21-40","41-60","61-80","81-100","101-120","121-140","141-160","161-180","181-200","201+")))


# ggplot(data=gh_combined_timeseries,aes(years_until_dis_bin,climate_value))+
#   geom_boxplot()+
#   facet_wrap(~variable,scale="free_y")
```

## 4.0 VEP II North Study Area Site Data

Accessed from:

### 4.1 Import data

```{r}
vepii_sites<-read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/db_vepii_n.csv")%>%
  subset(tlkp.Site.Function_DescGen=="Habitation")%>%
  subset(N.roomblocks >= 1)%>%
  subset(recordtypedesc=="Habitation" | recordtypedesc == "Great House"| recordtypedesc=="Oversized pitstructure" | recordtypedesc == "Great Kiva"| recordtypedesc=="Multi-wall Structure" | recordtypedesc == "Tower-Kiva") %>%
  #subset(Dataquality != 3)%>%
  st_as_sf(.,coords=c("UTMEast","UTMNorth"),crs=26912)%>%
  st_transform(.,crs=4269)%>%
  st_intersection(.,study_extent)%>%
  mutate(UNIQ_ID = 1:nrow(.))%>%
  subset(tlkp.Site.Function_DescGen=="Habitation")%>%
  subset(N.roomblocks >= 1)
```

### 4.2 Synthesize climate data

#### 4.2.1 Measures of Climate Stress

```{r}

for(i in 1:nrow(vepii_sites)){
  site <- vepii_sites[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- vepii_sites[i,c("UNIQ_ID")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  site_var <- terra::extract(r_var,site,exact=T)

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("mean_precipitation","year"))
  
  std_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("std_precipitation","year"))
  
  mean_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("mean_precipitation_change","year"))

  std_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("std_precipitation_change","year"))

  
  summary <- merge(mean_precipitation,std_precipitation,by="year")%>%
    merge(.,mean_precipitation_change,by="year")%>%
    merge(.,std_precipitation_change,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_vep <- summary
  }else {
    summary_all_vep <- merge(summary_all_vep,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   std_precipitation,
   mean_precipitation_change,
   std_precipitation_change)

write.csv(summary_all_vep,"./OUTPUT/climatevariability_vep_habitations.csv")
```

### 4.3 Synthesize occupation timeline

```{r}
#| echo: false
vep_df <- read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS/db_vepii_n.csv")%>%
  subset(tlkp.Site.Function_DescGen=="Habitation")%>%
  subset(N.roomblocks >= 1)%>%
  subset(recordtypedesc=="Habitation" | recordtypedesc == "Great House"| recordtypedesc=="Oversized pitstructure" | recordtypedesc == "Great Kiva"| recordtypedesc=="Multi-wall Structure" | recordtypedesc == "Tower-Kiva") %>%
  #subset(Dataquality != 3)%>%
  select(c("ph6","ph7","ph8","ph9","ph10","ph11","ph12",
           "ph13","ph14","ph15","ph16","ph17","ph18","ph19"))%>%
  mutate(UNIQ_ID = 1:nrow(.))%>%
  gather(phase,value,-UNIQ_ID)%>%
  mutate(value = ifelse(value<0,0,value))

vep_dates <- read.csv("./DATA/VEP_dates.csv")

vep_occupation_timeseries<- left_join(vep_df,vep_dates,c("phase"="PHASE"))%>%
  arrange(.,desc(UNIQ_ID),BEGIN_D)%>%
  subset(value>0)%>%
  select(-c("phase","value"))%>%
  setNames(c("ID","BEGIN_DATE","END_DATE"))%>%
  rowwise() %>%
  st_drop_geometry()%>%
  mutate(year = list(BEGIN_DATE:END_DATE)) %>%
  unnest(year) %>%
  select(c("ID", "year"))%>%
  group_by(ID)%>%
  mutate(year_until = find_years_until_disoccupation(year)-1)%>%
  setNames(c("ID","year","years_until_disoccupation"))

write.csv(vep_occupation_timeseries,"./OUTPUT/vep_occupation_timeseries.csv")
rm(vep_dates,vep_df)
```

### 4.4 Combine occupation and climate data

```{r}
#| echo: false

summary_all_vep <- read.csv("./OUTPUT/climatevariability_vep_habitations.csv")%>%select(-c("X"))

vep_occupation_timeseries <- read.csv("./OUTPUT/vep_occupation_timeseries.csv")%>%select(-c("X"))

vep_climate_timeseries<-summary_all_vep%>%
  gather(key=ID,value=climate_value,-c("year","variable"))%>%
  mutate(ID=as.numeric(sub("^X","",ID)))

vep_combined_timeseries <- merge(vep_climate_timeseries,vep_occupation_timeseries,by=c("ID","year"))

```

## 5.0 Combine and analyze

```{r}
#| echo: false
master_output <- rbind(gh_combined_timeseries,vep_combined_timeseries)

write.csv(master_output,"./OUTPUT/master_output.csv")
```

### 5.1 Broad correlations

Calculate correlations in these metrics.

```{r}
#| echo: false
master_output <- read.csv("./OUTPUT/master_output.csv")%>%
  select(-c("X"))

# mean(master_output$years_until_disoccupation)
# sd(master_output$years_until_disoccupation)
```

```{r}
#| echo: false
# head_ID <- master_output%>%
#   distinct(ID)%>%
#   slice_head(n=2)%>%
#   pull(ID)
# 
# tail_ID <- master_output%>%
#   distinct(ID)%>%
#   slice_tail(n=2)%>%
#   pull(ID)
# 
# ID_list <- c(head_ID,tail_ID)
# 
# master_sub <- master_output%>%
#   filter(ID %in% ID_list)

master_summary <- master_output%>%
  group_by(variable,years_until_disoccupation)%>%
  summarize(mean_value = mean(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_disoccupation>10 & years_until_disoccupation <670)

master_95 <- master_output%>%
  group_by(variable,years_until_disoccupation) %>%
  filter(climate_value >= quantile(climate_value, 0.95, na.rm = TRUE))%>%
  summarize(mean_value = min(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_disoccupation>10 & years_until_disoccupation <670)

master_05 <- master_output%>%
  group_by(variable,years_until_disoccupation) %>%
  filter(climate_value <= quantile(climate_value, 0.05, na.rm = TRUE))%>%
  summarize(mean_value = max(climate_value,na.rm=T))%>%
  ungroup()%>%
  mutate(mean_value=rollmean(mean_value,k=10,fill=NA,align="center"))%>%
  filter(years_until_disoccupation>10 & years_until_disoccupation <670)

  
png('./FIGURES/all_variables.png',height=18000,width=28000,res=1200)
ggplot(data=master_output,aes(years_until_disoccupation,climate_value))+
  geom_line(aes(group=ID),alpha=0.01,linewidth=2,col="#8AB17D")+
  geom_line(data=master_summary,aes(years_until_disoccupation,mean_value),color="grey20",linewidth=0.45)+
    geom_line(data=master_95,aes(years_until_disoccupation,mean_value),color="grey50",linewidth=0.25)+
    geom_line(data=master_05,aes(years_until_disoccupation,mean_value),color="grey50",linewidth=0.25)+
  labs(x="Years before Resettlement",
       y="Climate Variable (mm)",
       title=NULL)+
    theme_minimal()+
  facet_wrap(~variable,scale="free_y",labeller=as_labeller(c(`mean_precipitation` = "Mean Annual Precipitation",`mean_precipitation_change` = "Year-to-Year Change in Mean Precipitation (Temporal Variability)",`std_precipitation`="Standard Deviation in Annual Precipitation (Spatial Variability)",`std_precipitation_change`="Year-to-Year Change in Standard Deviation Precipitation (Spatio-Temporal Variability)")))+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        strip.text=element_text(size=15),
        axis.title = element_text(size=16,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10))
dev.off()


```

### 5.2 Outlying variability and outlying aridity

Calculate correlations among those things.

```{r}
#| echo: false

master_mean <- subset(master_output, variable == "mean_precipitation")
master_std <- subset(master_output, variable == "std_precipitation")
change_mean <- subset(master_output, variable == "mean_precipitation_change")
change_std <- subset(master_output, variable == "std_precipitation_change")


mean_df <- as.data.frame(matrix(0,99,4))%>%
  setNames(c("percentile","cor_coef","p_value","site_count"))%>%
  mutate(percentile = seq(0.01, 0.99, by = 0.01))

std_df <- as.data.frame(matrix(0,99,4))%>%
  setNames(c("percentile","cor_coef","p_value","site_count"))%>%
  mutate(percentile = seq(0.01, 0.99, by = 0.01))

mean_change_df <- as.data.frame(matrix(0,99,4))%>%
  setNames(c("percentile","cor_coef","p_value","site_count"))%>%
  mutate(percentile = seq(0.01, 0.99, by = 0.01))

std_change_df <- as.data.frame(matrix(0,99,4))%>%
  setNames(c("percentile","cor_coef","p_value","site_count"))%>%
  mutate(percentile = seq(0.01, 0.99, by = 0.01))


for(i in 1:99){
  perc <- mean_df[i,1] # for measuring amount of precip/negative cahnge (means)
  
  if(i <= 49){
  # mean precip
  mean_sub <- master_mean%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_s_test <- cor.test(mean_sub$years_until_disoccupation,mean_sub$climate_value)
  
  mean_df[i,c("cor_coef")] <- m_s_test$estimate
  mean_df[i,c("p_value")] <- m_s_test$p.value
  mean_df[i,c("site_count")] <- length(unique(mean_sub$ID))
  
  # std precip
  std_sub <- master_std%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_s_test <- cor.test(std_sub$years_until_disoccupation,std_sub$climate_value)
  
  std_df[i,c("cor_coef")] <- s_s_test$estimate
  std_df[i,c("p_value")] <- s_s_test$p.value
  std_df[i,c("site_count")] <- length(unique(std_sub$ID))
  
  # mean precip change
  mean_change_sub <- change_mean%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_c_test <- cor.test(mean_change_sub$years_until_disoccupation,mean_change_sub$climate_value)
  
  mean_change_df[i,c("cor_coef")] <- m_c_test$estimate
  mean_change_df[i,c("p_value")] <- m_c_test$p.value
  mean_change_df[i,c("site_count")] <- length(unique(mean_change_sub$ID))
  
  # std precip
  std_change_sub <- change_std%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value <= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_c_test <- cor.test(std_change_sub$years_until_disoccupation,std_change_sub$climate_value)
  
  std_change_df[i,c("cor_coef")] <- s_c_test$estimate
  std_change_df[i,c("p_value")] <- s_c_test$p.value
  std_change_df[i,c("site_count")] <- length(unique(std_change_sub$ID))
  
  } else {
    # mean precip
  mean_sub <- master_mean%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_s_test <- cor.test(mean_sub$years_until_disoccupation,mean_sub$climate_value)
  
  mean_df[i,c("cor_coef")] <- m_s_test$estimate
  mean_df[i,c("p_value")] <- m_s_test$p.value
  mean_df[i,c("site_count")] <- length(unique(mean_sub$ID))
  
  # std precip
  std_sub <- master_std%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_s_test <- cor.test(std_sub$years_until_disoccupation,std_sub$climate_value)
  
  std_df[i,c("cor_coef")] <- s_s_test$estimate
  std_df[i,c("p_value")] <- s_s_test$p.value
  std_df[i,c("site_count")] <- length(unique(std_sub$ID))
  
  # mean precip change
  mean_change_sub <- change_mean%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  m_c_test <- cor.test(mean_change_sub$years_until_disoccupation,mean_change_sub$climate_value)
  
  mean_change_df[i,c("cor_coef")] <- m_c_test$estimate
  mean_change_df[i,c("p_value")] <- m_c_test$p.value
  mean_change_df[i,c("site_count")] <- length(unique(mean_change_sub$ID))
  
  # std precip
  std_change_sub <- change_std%>%
    group_by(years_until_disoccupation) %>%
    filter(climate_value >= quantile(climate_value, perc, na.rm = TRUE)) %>%
    ungroup()
  
  s_c_test <- cor.test(std_change_sub$years_until_disoccupation,std_change_sub$climate_value)
  
  std_change_df[i,c("cor_coef")] <- s_c_test$estimate
  std_change_df[i,c("p_value")] <- s_c_test$p.value
  std_change_df[i,c("site_count")] <- length(unique(std_change_sub$ID))
  }

}

mean_df<- mean_df%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))
         
std_df<- std_df%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))

mean_change_df<- mean_change_df%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))
         
std_change_df<- std_change_df%>%
  mutate(sig = ifelse(p_value <= 0.05,1,0))

         
mean_df_sig <- mean_df%>%
  filter(sig == 1)
std_df_sig <- std_df%>%
  filter(sig == 1)
mean_change_df_sig <- mean_change_df%>%
  filter(sig == 1)
std_change_df_sig <- std_change_df%>%
  filter(sig == 1)

```

```{r}
  # change_mean_sub <- change_mean%>%
  #   group_by(years_until_disoccupation) %>%
  #   filter(climate_value >= quantile(climate_value, 0.99, na.rm = TRUE)) %>%
  #   ungroup()
  # 
  # ggplot(data=change_mean_sub,aes(years_until_disoccupation,climate_value))+
  #   geom_point()
```

```{r}
#| echo: false

p1 <- ggplot()+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill=NA,color="#4D4D4D80")+
  geom_line(data=mean_df,aes(percentile,cor_coef),col="#bb3e03",lwd=1.2)+
  geom_point(data=mean_df_sig,aes(percentile,cor_coef),col="#bb3e03",size=1.5)+
  geom_line(data=mean_change_df,aes(percentile,cor_coef),col="#ee9b00",lwd=1.2)+
  geom_point(data=mean_change_df_sig,aes(percentile,cor_coef),col="#ee9b00",size=1.5)+
  
   annotate("text",x=0.48,y=0.85,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=1,lineheight = 0.8)+ 
    annotate("text",x=0.49,y=0.73,label="Low Precip.",size=4, color="#bb3e03",hjust=1)+
   annotate("text",x=0.48,y=0.67,label="Negative Temporal Variability",size=4, color="#ee9b00",hjust=1)+

  annotate("text",x=0.52,y=-0.7,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=0,lineheight = 0.8)+ 
  annotate("text",x=0.52,y=-0.82,label="High Precip.",size=4, color="#bb3e03",hjust=0)+
  annotate("text",x=0.52,y=-0.88,label="Positive Temporal Variability",size=4, color="#ee9b00",hjust=0)+


  labs(x=NULL,
       y="Coorelation Coefficient",
       title= "A")+
  
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=20)),
        plot.margin = margin(10,10,10,10),
        title=element_text(size=17))

p2<- ggplot()+
  annotate("rect",xmin=0.01,xmax=0.5,ymin=0,ymax=1,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill="grey70",alpha=0.1)+
  annotate("rect",xmin=0.5,xmax=0.99,ymin=-1,ymax=0,fill=NA,color="#4D4D4D80")+
  geom_line(data=std_df,aes(percentile,cor_coef),col="#005f73",lwd=1.2)+
  geom_point(data=std_df_sig,aes(percentile,cor_coef),col="#005f73",size=1.5)+
  geom_line(data=std_change_df,aes(percentile,cor_coef),col="#94d2bd",lwd=1.2)+
  geom_point(data=std_change_df_sig,aes(percentile,cor_coef),col="#94d2bd",size=1.5)+
  
     annotate("text",x=0.48,y=0.85,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=1,lineheight = 0.8)+ 
  annotate("text",x=0.48,y=0.73,label="Less Spatial Variability",size=4, color="#005f73",hjust=1)+
      annotate("text",x=0.48,y=0.67,label="Less Spatio-temporal Variability",size=4, color="#94d2bd",hjust=1)+
  
    annotate("text",x=0.52,y=-0.7,label="Closer to \nResettlement \nAssociated with:",size=4, color="grey10",hjust=0,lineheight = 0.8)+ 
  annotate("text",x=0.52,y=-0.82,label="More Spatial Variability",size=4, color="#005f73",hjust=0)+
  annotate("text",x=0.52,y=-0.88,label="More Spatio-temporal Variability",size=4, color="#94d2bd",hjust=0)+
  
  labs(x=NULL,
       y=NULL,
       title= "B")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        plot.margin = margin(10,10,10,10),
        title=element_text(size=17))
p3 <- ggplot()+
  geom_point(data=mean_change_df,aes(percentile, site_count),col="#ee9b00")+
  geom_point(data=mean_df,aes(percentile, site_count),col="#bb3e03")+
  labs(x="Percentile",
       y="No. of Sites",
       title= "C")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(0,6000),breaks=c(0,2000,4000,6000))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.y = element_text(size=15,margin=margin(r=20)),
        axis.title.x = element_text(size=15,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10),
        title=element_text(size=17))

p4<-ggplot()+
  geom_point(data=std_df,aes(percentile, site_count),col="#005f73")+
  geom_point(data=std_change_df,aes(percentile, site_count),col="#94d2bd")+
  labs(x="Percentile",
       y=NULL,
       title= "D")+
  scale_x_continuous(breaks = c(0.01,0.25,0.5,0.75,0.99), labels = c("<0.01","<0.25","0.5",">0.75",">0.99"))+
  scale_y_continuous(limits=c(0,6000),breaks=c(0,2000,4000,6000))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=10),
        axis.title.x = element_text(size=15,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10),
        title=element_text(size=17))

png('./FIGURES/standarddeviation_and_change.png',height=16000,width=24000,res=1200)
ggarrange(p1,p2,p3,p4,
          ncol=2,nrow=2,align="hv",heights=c(1.5,0.7))
dev.off()

```

```{r}
# test <- master_output%>%
#   spread(key=variable,climate_value)


mod <- lm(years_until_disoccupation ~ mean_precipitation,data=test)
mod_s <- summary(mod)
mod_s$sigma

mod1 <- lm(years_until_disoccupation ~ mean_precipitation_change,data=test)
mod1_s <- summary(mod1)
mod1_s$sigma

mod2 <- lm(years_until_disoccupation ~ std_precipitation,data=test)
mod2_s <- summary(mod2)
mod2_s$sigma

mod3 <- lm(years_until_disoccupation ~ std_precipitation_change,data=test)
mod3_s <- summary(mod3)
mod3_s$sigma


mod4 <- lm(years_until_disoccupation ~ mean_precipitation+mean_precipitation_change+ std_precipitation+std_precipitation_change,data=test)
mod4_s <- summary(mod4)
mod4_s$sigma

mod5 <- lm(years_until_disoccupation ~ mean_precipitation_change+ std_precipitation+std_precipitation_change,data=test)
mod5_s <- summary(mod5)
mod5_s$sigma

```

## 6.0 Dendrochronological Data

Accessed from ....:

1.  copy formatting from above

### 6.1 Import data

```{r}
#| echo: false
dendro_dates <- read.csv("C:/Users/sfield3/Documents/RESEARCH/DATA GENERAL/SENSITIVE SITE LOCATIONS//BOCINSKY_ET_AL_2015_TREE_RINGS.csv")%>%
  st_as_sf(.,coords=c("LON","LAT"),crs=4269)%>%
  st_intersection(.,study_extent)%>%
  subset(.,Outer_Symbol == "V"|Outer_Symbol=="RLB"|Outer_Symbol=="R"|Outer_Symbol=="+B"|Outer_Symbol=="++G"|Outer_Symbol=="RG"|Outer_Symbol=="+V"|Outer_Symbol=="L"|Outer_Symbol=="+RLB"|Outer_Symbol=="CB"|Outer_Symbol=="+L"|Outer_Symbol=="+LB"|Outer_Symbol=="LGB"|Outer_Symbol=="C"|Outer_Symbol=="CL"|Outer_Symbol=="+R"|Outer_Symbol=="+CL"|Outer_Symbol=="CGB"|Outer_Symbol=="RLG"|Outer_Symbol=="GB"|Outer_Symbol=="VL"|Outer_Symbol=="GLB"|Outer_Symbol=="+GLB"|Outer_Symbol=="BB"|Outer_Symbol=="+LG"|Outer_Symbol=="+RGB"|Outer_Symbol=="CLB"|Outer_Symbol=="c"|Outer_Symbol=="r"|Outer_Symbol=="cB"|Outer_Symbol=="+r"|Outer_Symbol=="rB"|Outer_Symbol=="+VL"|Outer_Symbol=="CG"|Outer_Symbol=="+CG"|Outer_Symbol=="RC"|Outer_Symbol=="LB "|Outer_Symbol=="rGB"|Outer_Symbol== "RB COM"|Outer_Symbol== "R COMP"|Outer_Symbol== "R INC")%>%
  subset(.,!is.na(Site_Number))%>%
  mutate(ID=1:nrow(.))


duplicates <- st_coordinates(dendro_dates)%>%
  duplicated()

dendro_dates_unique_loc <- dendro_dates[!duplicates,]

ggplot()+
  geom_sf(data=dendro_dates)
```

### 6.2 Synthesize Climate Data

```{r}
#| echo: false

for(i in 1:nrow(dendro_dates_unique_loc)){
  site <- dendro_dates_unique_loc[i,]%>%
    st_buffer(.,2000)%>%
    st_transform(.,crs=4326)

  site_name <- dendro_dates_unique_loc[i,c("ID")]%>%
                st_drop_geometry()%>%
                as.character()

  site_precip <- terra::extract(r,site,exact=T)
  
  site_var <- terra::extract(r_var,site,exact=T)

  mean_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("mean_precipitation","year"))
  
  std_precipitation <- site_precip%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=1:2000)%>%
    setNames(c("std_precipitation","year"))
  
  mean_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(mean,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("mean_precipitation_change","year"))

  std_precipitation_change <- site_var%>%
    select(-c("ID","fraction"))%>%
    summarise_all(sd,na.rm=T)%>%
    t()%>%
    as.data.frame()%>%
    mutate(year=2:2000)%>%
    setNames(c("std_precipitation_change","year"))

  
  summary <- merge(mean_precipitation,std_precipitation,by="year")%>%
    merge(.,mean_precipitation_change,by="year")%>%
    merge(.,std_precipitation_change,by="year")%>%
    gather(.,key="variable",value="value",-year)%>%
    setNames(c("year","variable",site_name))

  if (i == 1){
    summary_all_dendro <- summary
  }else {
    summary_all_dendro <- merge(summary_all_dendro,summary, by=c("year","variable"))
  }
}

rm(site_precip,
   site_var,
   mean_precipitation,
   std_precipitation,
   mean_precipitation_change,
   std_precipitation_change)

write.csv(summary_all_dendro,"./OUTPUT/climatevariability_dendro.csv")

```

### 6.3 Synthesize occupation timeline

```{r}
#| echo: false
dendro_df <-  st_drop_geometry(dendro_dates)%>%
  select(c("ID","Outer_Date_AD"))%>%
  setNames(c("ID","year"))%>%
  mutate(count= 1)%>%
  group_by(year)%>%
  reframe(count=sum(count))%>%
  filter(., year >= 1033 & year <= 1335)%>%
  complete(year = 1033:1335)%>%
  mutate(interpolated = ifelse(is.na(count),1,0),
         count_raw = count,
         count = zoo::na.approx(count))


png('./FIGURES/TreeRing_timeseries.png',height=11000,width=18000,res=1200)
ggplot()+
  geom_line(data=dendro_df,aes(year,count,group=1,color=as.factor(interpolated)),linewidth=1)+
  scale_color_manual(values=c("#8AB17D","#EE8959"))+
  labs(title="Number of Tree Ring cutting Dates by Year",
       x="Year",
       y="Number of Cutting Dates",
       subtitle="orange line represents interpolated values",
  )+
  scale_x_continuous(limits=c(1000,1350),breaks=c(1000,1100,1200,1300))+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        plot.title = element_text(size=17,margin=margin(t=10,b=3)),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=20)),
        axis.title.x = element_text(size=15,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10),
        strip.text.x = element_text(size=13,margin=margin(b=10,t=10)),
        legend.position = "none",
        plot.subtitle = element_text(size=12,margin=margin(t=5,b=3,r=100)))
dev.off()


write.csv(dendro_df,"./OUTPUT/dendro_activity_timeseries.csv")
```

### 6.4 Combine Climate and Occupation

Used min and max because we want to find the values that are characteristic of the respective percentile limit.

```{r}
#| echo: false
summary_all_dendro <- read.csv("./OUTPUT/climatevariability_dendro.csv")%>%
  select(-c("X"))%>%
  filter(., year >= 1033 & year <= 1335)%>%
  gather(key=ID,value=climate_value,-c("year","variable"))%>%
  mutate(ID=as.numeric(sub("^X","",ID)))

dendro_mean <- subset(summary_all_dendro, variable == "mean_precipitation")%>%
    group_by(year) %>%
    filter(climate_value <= quantile(climate_value, 0.05, na.rm = TRUE))%>%
    summarise(mean_precipitation=max(climate_value,na.rm=T))
    

dendro_std <- subset(summary_all_dendro, variable == "std_precipitation")%>%
    group_by(year) %>%
    filter(climate_value >= quantile(climate_value, 0.95, na.rm = TRUE))%>%
    summarise(std_precipitation=min(climate_value,na.rm=T))

dendro_change_mean <- subset(summary_all_dendro, variable == "mean_precipitation_change")%>%
    group_by(year) %>%
    filter(climate_value <= quantile(climate_value, 0.05, na.rm = TRUE))%>%
    summarise(mean_precipitation_change=max(climate_value,na.rm=T))

dendro_change_std <- subset(summary_all_dendro, variable == "std_precipitation_change")%>%
    group_by(year) %>%
    filter(climate_value >= quantile(climate_value, 0.95, na.rm = TRUE))%>%
    summarise(std_precipitation_change=min(climate_value,na.rm=T))

dendro_timeseries <- read.csv("./OUTPUT/dendro_activity_timeseries.csv")%>%select(-c("X"))


dendro_combined_timeseries <- merge(dendro_timeseries,dendro_mean,by=c("year"))%>%
merge(.,dendro_std,by=c("year"))%>%
merge(.,dendro_change_mean,by=c("year"))%>%
merge(.,dendro_change_std,by=c("year"))

             
             
# ggplot()+
#   geom_line(data=dendro_combined_timeseries,aes(year,mean_precipitation),col="#bb3e03")+
#   geom_line(data=dendro_combined_timeseries,aes(year,std_precipitation),col="#005f73")+
#   geom_line(data=dendro_combined_timeseries,aes(year,mean_precipitation_change),col="#ee9b00")+
#   geom_line(data=dendro_combined_timeseries,aes(year,std_precipitation_change),col="#94d2bd")



# p < 0.05 = stationary so good for granger causality
adf.test(dendro_combined_timeseries[,2])
adf.test(dendro_combined_timeseries[,5])
adf.test(dendro_combined_timeseries[,6])
adf.test(dendro_combined_timeseries[,7])
adf.test(dendro_combined_timeseries[,8])
```

```{r}
#| echo: false

gc_results <- data.frame(year_lag=1:25,
                         mean_precipitation=NA,
                         mean_precipitation_change=NA,
                         std_precipitation=NA,
                         std_precipitation_change=NA)

for(i in c("mean_precipitation","std_precipitation","mean_precipitation_change","std_precipitation_change")){
  
  dendro_sample <- dendro_combined_timeseries%>%
    select(all_of(c("count",i)))
  
  for(j in 1:25){
      gt<-grangertest(dendro_sample[,1] ~ dendro_sample[,2], order = j)
      gc_results[j,c(i)] <- gt$`Pr(>F)`[2]
  }
}
```

```{r}
png('./FIGURES/GrangerCausality.png',height=11000,width=18000,res=1200)
ggplot()+
  geom_line(data=gc_results,aes(year_lag,mean_precipitation),col="#bb3e03")+
  geom_point(data=gc_results,aes(year_lag,mean_precipitation),col="#bb3e03")+
  geom_line(data=gc_results,aes(year_lag,std_precipitation),col="#005f73")+
  geom_point(data=gc_results,aes(year_lag,std_precipitation),col="#005f73")+
  geom_line(data=gc_results,aes(year_lag,mean_precipitation_change),col="#ee9b00")+
  geom_point(data=gc_results,aes(year_lag,mean_precipitation_change),col="#ee9b00")+
  geom_line(data=gc_results,aes(year_lag,std_precipitation_change),col="#94d2bd")+
  geom_point(data=gc_results,aes(year_lag,std_precipitation_change),col="#94d2bd")+
  geom_hline(yintercept=0.05,col="grey30",linetype="dotted")+
  labs(x="Year Lag",
       y="Granger Causality P-value",
       title=NULL)+
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.1, color = "grey75"),
        axis.text = element_text(size=12),
        axis.title.y = element_text(size=15,margin=margin(r=20)),
        axis.title.x = element_text(size=15,margin=margin(t=20)),
        plot.margin = margin(10,10,10,10),
        legend.title=element_text(size=15),
        legend.text = element_text(size=12))
dev.off()
```
